---
title: 'Demultiplexing the PBMC pool: PP1'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(DropletUtils)
library(here)
library(ggplot2)
library(Seurat)
library(cowplot)
library(patchwork)
library(scater)
library(dplyr)
library(vcfR)
library(janitor)
library(stringr)
})
set.seed(1990)

```

# 2 Set names for this pool

```{r}
# Specify batch name
batch_name <- "PP1"
num_of_captures <- 3

# Specify capture name
capture_names <- c(paste0(batch_name, "-c",1:num_of_captures))
capture_names <- setNames(capture_names, capture_names)

# Assign sample ID to HTO ID
# Manually list sample names matching HTO id (in alphabetical order)
samples <- c("M1C198B", #HTO3
             "M1C208A", #HTO6
             "M1C166(1)", #HTO7
             "M1C199B", #HTO8
             "M1C207B", #HTO10
             "M1C190C", #HTO12
             "M1N087", #HTO13
             "M1N092" #HTO14
)

```

# 3 Read CellBender outputs
Ambient RNA and empty droplets were removed by CellBender (see code/cellbender.sh).
```{r}
captures <- setNames(
  here("data",
       "cellbender",
       capture_names,
       paste0(capture_names,".cellbender_filtered.h5")),
  capture_names)

sce <- read10xCounts(samples = captures, col.names = TRUE)
stopifnot(!anyDuplicated(colnames(sce)))

sce <- splitAltExps(
  sce,
  rowData(sce)$Type,
  "Gene Expression")

# Tidy up colData
sce$Capture <- factor(sce$Sample)
capture_names <- levels(sce$Capture)
capture_names <- setNames(capture_names, capture_names)
sce$Sample <- NULL

# Number of droplets after CellBender
dim(sce)

```

# 4 HTO demultiplexing

HTODemux from the Seurat package was used to demultiplex droplets. 
Demultiplexing was performed separately on each capture.

## 4.1 Prepare HTO data
```{r}
# change DelayedMatrix to dgCMatrix
assay(sce, "counts") <- as(assay(sce, "counts"), "dgCMatrix")

# Read features.csv used in cellranger multi
features <- read.csv(here("data",
                          "sample_sheets",
                          paste0(batch_name,".features.csv")))

# HTO id are always in alphabetical order
features

# Rename assay
is_hto <- grepl("^Human_HTO", rownames(altExp(sce, "Antibody Capture")))
altExp(sce, "HTO") <- altExp(sce, "Antibody Capture")[is_hto, ]
altExp(sce, "Antibody Capture") <- NULL

```

## 4.2 HTO Demultiplexing
```{r}
# load in umi and hto matrix
umis <- counts(sce)
htos <- counts(altExp(sce))

# Select cell barcodes detected by both RNA and HTO
joint.bcs <- intersect(colnames(umis), colnames(htos))

# Subset RNA and HTO counts by joint cell barcodes
umis <- umis[, joint.bcs]
htos <- as.matrix(htos[, joint.bcs])

# Confirm that the HTO have the correct names
rownames(htos)

# Perform demultiplexing separately on each capture
obj.list <- list()

lappend <- function (lst, ...){
  lst <- c(lst, list(...))
  return(lst)
}

for (cn in capture_names) {
  message(cn)
  umi = as.matrix(umis[, sce$Capture == cn])
  hto = as.matrix(htos[, sce$Capture == cn])
  
  # Setup Seurat object
  hashtag <- CreateSeuratObject(counts = umi)
  
  # Normalize RNA data with log normalization
  hashtag <- NormalizeData(hashtag)
  
  # Find and scale variable features
  hashtag <- FindVariableFeatures(hashtag, selection.method = "mean.var.plot")
  hashtag <- ScaleData(hashtag, features = VariableFeatures(hashtag))
  
  # Add HTO data as a new assay independent from RNA
  hashtag[["HTO"]] <- CreateAssayObject(counts = hto)
  
  # Normalize HTO data, here we use centered log-ratio (CLR) transformation
  hashtag <- NormalizeData(hashtag, assay = "HTO", normalization.method = "CLR")
  
  # Demultiplex cells based on HTO enrichment
  hashtag <- HTODemux(hashtag, assay = "HTO", positive.quantile = 0.99)
  
  # Rename hash.ID column 
  hashtag$hash.ID <- factor(gsub("-","_",hashtag$hash.ID), 
                            levels = c(features$id, "Doublet", "Negative"))
  obj.list <- lappend(obj.list, hashtag)
}

names(obj.list) <- capture_names
hashtag <- NULL

```

## 4.3 Add HTO demultiplexing results to SCE
```{r}
# add hash.ID column to SCE object
hash.ID <- Reduce(c,
                  lapply(obj.list, function(x) 
                    setNames(x@meta.data$hash.ID, colnames(x))))
sce$HTODemux.ID <- hash.ID

# Store HTODemux results to SCE object
sce$HTODemux_result <- bind_rows(lapply(obj.list, function(x) x@meta.data))

# Add sample ID to SCE object
sampleID.HTO <- setNames(c(samples,"Doublet","Negative"), levels(sce$HTODemux.ID))
sampleID.HTO <- sampleID.HTO[sce$HTODemux.ID]
names(sampleID.HTO) <- colnames(sce)
sce$sampleID.HTO <- factor(sampleID.HTO, levels=c(samples,"Doublet","Negative"))

```

# 5 Visualize HTO demultiplexing results

```{r}
# add colour
sce$colours <- S4Vectors::make_zero_col_DFrame(ncol(sce))

hto_colours <- setNames(
  c(palette.colors(nlevels(sce$HTODemux.ID)-2, "Tableau 10"),"#ced4da","#6c757d"),
  levels(sce$HTODemux.ID))
sce$colours$hto_colours <- hto_colours[sce$HTODemux.ID]

sample_colours <- setNames(
  c(palette.colors(nlevels(sce$sampleID.HTO)-2, "Tableau 10"),"#ced4da","#6c757d"),
  levels(sce$sampleID.HTO))
sce$colours$sample_colours <- sample_colours[sce$sampleID.HTO]

capture_colours <- setNames(
  palette.colors(nlevels(sce$Capture), "Accent"),
  levels(sce$Capture))
sce$colours$capture_colours <- capture_colours[sce$Capture]

```

## 5.1 Number of singlets
::::: {.panelset}

::: {.panel}

### PP1-c1 {.panel-name}

```{r}
table(obj.list$`PP1-c1`$HTO_classification.global)

```

:::

::: {.panel}

### PP1-c2 {.panel-name}

```{r}
table(obj.list$`PP1-c2`$HTO_classification.global)

```

:::

::: {.panel}

### PP1-c3 {.panel-name}

```{r}
table(obj.list$`PP1-c3`$HTO_classification.global)

```

:::

:::::

## 5.2 HTO expression in ridge plot
::::: {.panelset}

::: {.panel}

### PP1-c1 {.panel-name}

```{r fig.width=24,fig.height=12}
# Group cells based on the max HTO signal
Idents(obj.list$`PP1-c1`) <- "hash.ID"
RidgePlot(obj.list$`PP1-c1`, assay = "HTO", 
          features = rownames(obj.list$`PP1-c1`[["HTO"]]))

```

:::

::: {.panel}

### PP1-c2 {.panel-name}

```{r fig.width=24,fig.height=12}
# Group cells based on the max HTO signal
Idents(obj.list$`PP1-c2`) <- "hash.ID"
RidgePlot(obj.list$`PP1-c2`, assay = "HTO", 
          features = rownames(obj.list$`PP1-c2`[["HTO"]]))

```

:::

::: {.panel}

### PP1-c3 {.panel-name}

```{r fig.width=24,fig.height=12}
# Group cells based on the max HTO signal
Idents(obj.list$`PP1-c3`) <- "hash.ID"
RidgePlot(obj.list$`PP1-c3`, assay = "HTO", 
          features = rownames(obj.list$`PP1-c3`[["HTO"]]))

```

:::

:::::

## 5.3 HTO expression in heatmap
::::: {.panelset}

::: {.panel}

### PP1-c1 {.panel-name}

```{r fig.width=8,fig.height=6}
HTOHeatmap(obj.list$`PP1-c1`, assay = "HTO", ncells = 5000) + 
  ggtitle("PP1-c1") + 
  theme(plot.title = element_text(hjust=0.5, size=10))

```
:::

::: {.panel}

### PP1-c2 {.panel-name}

```{r fig.width=8,fig.height=6}
HTOHeatmap(obj.list$`PP1-c2`, assay = "HTO", ncells = 5000) + 
  ggtitle("PP1-c2") + 
  theme(plot.title = element_text(hjust=0.5, size=12))

```
:::

::: {.panel}

### PP1-c3 {.panel-name}

```{r fig.width=8,fig.height=6}
HTOHeatmap(obj.list$`PP1-c3`, assay = "HTO", ncells = 5000) + 
  ggtitle("PP1-c3") + 
  theme(plot.title = element_text(hjust=0.5, size=12))

```

:::

:::::

## 5.4 tSNE embeddings for HTOs
::::: {.panelset}

::: {.panel}

### PP1-c1 {.panel-name}

```{r}
# remove negative cells from the object
hashtag.subset <- subset(obj.list$`PP1-c1`, idents = "Negative", invert = TRUE)

# calculate a tSNE embedding of the HTO data
DefaultAssay(hashtag.subset) <- "HTO"
hashtag.subset <- ScaleData(hashtag.subset, features = rownames(hashtag.subset))
hashtag.subset <- RunPCA(hashtag.subset, features = rownames(hashtag.subset), approx = FALSE)
hashtag.subset <- RunTSNE(hashtag.subset, dims = 1:length(rownames(hashtag.subset)), 
                          perplexity = 100, check_duplicates = FALSE)

# tSNE plot
DimPlot(hashtag.subset)

```

:::

::: {.panel}

### PP1-c2 {.panel-name}

```{r}
# remove negative cells from the object
hashtag.subset <- subset(obj.list$`PP1-c2`, idents = "Negative", invert = TRUE)

# calculate a tSNE embedding of the HTO data
DefaultAssay(hashtag.subset) <- "HTO"
hashtag.subset <- ScaleData(hashtag.subset, features = rownames(hashtag.subset))
hashtag.subset <- RunPCA(hashtag.subset, features = rownames(hashtag.subset), approx = FALSE)
hashtag.subset <- RunTSNE(hashtag.subset, dims = 1:length(rownames(hashtag.subset)), 
                          perplexity = 100, check_duplicates = FALSE)

# tSNE plot
DimPlot(hashtag.subset)

```

:::

::: {.panel}

### PP1-c3 {.panel-name}

```{r}
# remove negative cells from the object
hashtag.subset <- subset(obj.list$`PP1-c3`, idents = "Negative", invert = TRUE)

# calculate a tSNE embedding of the HTO data
DefaultAssay(hashtag.subset) <- "HTO"
hashtag.subset <- ScaleData(hashtag.subset, features = rownames(hashtag.subset))
hashtag.subset <- RunPCA(hashtag.subset, features = rownames(hashtag.subset), approx = FALSE)
hashtag.subset <- RunTSNE(hashtag.subset, dims = 1:length(rownames(hashtag.subset)), 
                          perplexity = 100, check_duplicates = FALSE)

# tSNE plot
DimPlot(hashtag.subset)

```

:::

:::::

## 5.5 Number of demultiplexed droplets
```{r}
df <- as.data.frame(colData(sce))
```
::::: {.panelset}

::: {.panel}

### Named by HTO {.panel-name}

```{r fig.width=10,fig.height=4}
# Number of droplets per HTO classification
p1 <- ggplot(df) + 
  geom_bar(
    aes(
      x = factor(hash.ID, levels(sce$HTODemux.ID)),
      fill=hash.ID), 
    position = position_stack(reverse = TRUE)) + 
  geom_text(stat='count', aes(x = hash.ID, label=..count..), hjust=1, size=2.5) +
  coord_flip() +
  scale_fill_manual(values = hto_colours) +
  ylab("Number of droplets") +
  xlab("HTODemux classification") +
  theme_cowplot(font_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 + NoLegend() + p1 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names)))

```

:::

::: {.panel}

### Named by sample ID {.panel-name}

```{r fig.width=10,fig.height=4}
# Number of droplets per sample
p2 <- ggplot(df) + 
  geom_bar(
    aes(
      x = factor(sampleID.HTO, levels(sce$sampleID.HTO)),
      fill=sampleID.HTO), 
    position = position_stack(reverse = TRUE)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), hjust=1, size=2.5) +
  coord_flip() +
  scale_fill_manual(values = sample_colours) +
  ylab("Number of droplets") +
  xlab("Sample ID") +
  theme_cowplot(font_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 + NoLegend() + p2 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names)))

```

:::

:::::

## 5.6 Proportion of singlets per capture
```{r fig.width=6,fig.height=3}
p3 <- ggplot(df) + 
  geom_bar(
    aes(
      x = Capture,
      fill= HTODemux_result.HTO_classification.global), 
    position = position_fill(reverse = TRUE)) + 
  coord_flip() +
  scale_fill_manual(values = setNames(c("#6c757d","#ced4da","#cfe1b9"),
                                      c("Negative","Doublet","Singlet"))) +
  ylab("Proportion of droplets") +
  xlab("HTODemux classification") +
  theme_cowplot(font_size = 8) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3 +
  guides(fill=guide_legend(title="HTO_classification.global"))


```


# 6 Genetic demultiplexing

SNPs were called from each capture using cellsnp-lite v1.2.3. 
Based on the SNP information, cells were assigned to genetic donors using vireo 
v0.5.8.

## 6.1 Read vireo results
```{r}
vireo_df <- do.call(
  rbind,
  lapply(capture_names, function(cn) {
    message(cn)
    vireo_df <- read.table(
      here("data", "vireo", cn, "donor_ids.tsv"),
      header = TRUE)
    # Rename 'doublet' and 'unassigned' to match the terms used in HTODemux 
    vireo_df$donor_id <- gsub("doublet", "Doublet", vireo_df$donor_id)
    vireo_df$donor_id <- gsub("unassigned", "Negative", vireo_df$donor_id)
    vireo_df$donor_id <- paste0(cn, "_", vireo_df$donor_id)
    captureNumber <- str_sub(cn, start= -1)
    vireo_df$colname <- paste0(captureNumber, "_", vireo_df$cell)
    # reorder columns to matches SCE.
    j <- match(colnames(sce)[sce$Capture == cn], vireo_df$colname)
    stopifnot(!anyNA(j))
    vireo_df <- vireo_df[j, ]
  }))

```

## 6.2 Summarise vireo results
```{r}
knitr::kable(
  tabyl(vireo_df, donor_id) %>%
    adorn_pct_formatting(1),
  caption = "Assignment of droplets to donors using vireo.")
```

## 6.3 Identify the best match between donors from the combination of captures
```{r}
# capture 1 is used as base
cn1 <- capture_names[1]

# read vcf
f1 <- here("data","vireo",cn1,"GT_donors.vireo.vcf.gz")
x1 <- read.vcfR(f1, verbose=FALSE)

# create unique ID for each locus in each capture.
y1 <- paste(
  x1@fix[,"CHROM"],
  x1@fix[,"POS"],
  x1@fix[,"REF"],
  x1@fix[,"ALT"],
  sep = "_")

# match donors of every remaining captures to capture 1
f.best_match_df <- data.frame()

for (cn2 in capture_names[2:length(capture_names)]) {
  
  # read vcf
  f2 <- here("data","vireo",cn2,"GT_donors.vireo.vcf.gz")
  x2 <- read.vcfR(f2, verbose=FALSE)
  
  # create unique ID for each locus in each capture.
  y2 <- paste(
    x2@fix[,"CHROM"],
    x2@fix[,"POS"],
    x2@fix[,"REF"],
    x2@fix[,"ALT"],
    sep = "_")
  
  # only keep the loci in common between the 2 captures.
  i1 <- na.omit(match(y2, y1))
  i2 <- na.omit(match(y1, y2))
  
  # construct genotype matrix at common loci from the 2 captures.
  donor_names <- colnames(x1@gt)[-1]
  
  g1 <- apply(
    x1@gt[i1, donor_names],
    2,
    function(x) sapply(strsplit(x, ":"), `[[`, 1))
  g2 <- apply(
    x2@gt[i2, donor_names],
    2,
    function(x) sapply(strsplit(x, ":"), `[[`, 1))
  
  # count number of genotype matches between pairs of donors (one from each 
  # capture) and convert to a proportion.
  z <- matrix(
    NA_real_,
    nrow = length(donor_names),
    ncol = length(donor_names),
    dimnames = list(donor_names, donor_names))
  
  for (i in rownames(z)) {
    for (j in colnames(z)) {
      z[i, j] <- sum(g1[, i] == g2[, j]) / nrow(g1)
    }
  }
  rownames(z) <- paste(cn1, rownames(z),sep="_")
  colnames(z) <- paste(cn2, colnames(z),sep="_")
  
  # look for the best match between donors
  best_match_df <- data.frame(
    cn1 = rownames(z),
    cn2 = apply(
      z, 1,
      function(x) colnames(z)[which.max(x)]),
    check.names = FALSE)
  
  knitr::kable(
    dplyr::select(best_match_df, everything()),
    caption = paste0("Best match between donors between ", 
                     cn1, " and ", cn2, "."),
    row.names = FALSE)
  
  stopifnot(identical(colnames(sce), vireo_df$colname))
  
  # visualize the best match in a heatmap
  pheatmap::pheatmap(
    z,
    color = viridisLite::inferno(101),
    cluster_rows = FALSE,
    cluster_cols = FALSE,
    main = paste0("Proportion of matching genotypes between ", 
                  cn1, " and ", cn2, "."))
  
  # add matching results to the final best match data frame
  if (nrow(f.best_match_df)==0) {
    f.best_match_df <- best_match_df
  } else {
    f.best_match_df <- f.best_match_df %>% 
      mutate(c=best_match_df$cn2)
    colnames(f.best_match_df) <- paste0("cn",1:ncol(f.best_match_df))
  }
}

# add genetic donor name
f.best_match_df$genetic_donor <- paste0("donor_", LETTERS[1:length(donor_names)])

knitr::kable(
  dplyr::select(f.best_match_df, genetic_donor, everything()),
  caption = "Best match between donors from the two captures.",
  row.names = FALSE)

# transform to a long data frame 
best_match_long_df <- tidyr::pivot_longer(
  data=f.best_match_df,
  cols=starts_with("cn")) %>%
  dplyr::select(genetic_donor, value)

```


## 6.4 Add genetic demultiplexing results to SCE

```{r}
# add genetic donor to SCE object
sce$vireo <- DataFrame(
  vireo_df[, setdiff(colnames(vireo_df), c("cell", "colname"))])

sce$genetic_donor <- left_join(
  vireo_df,
  best_match_long_df, 
  by = c("donor_id" = "value")) %>%
  mutate(
    genetic_donor = factor(
      case_when(
        is.na(genetic_donor) ~ sub(paste0(batch_name,"-c[0-9]_"), "", donor_id),
        TRUE ~ genetic_donor),
      levels = c(paste0("donor_", LETTERS[1:length(donor_names)]), "Doublet", "Negative"))) %>%
  pull(genetic_donor)

# generate a table of HTO to genetic droplets
tb <- as.data.frame.matrix(
  table(as.data.frame(colData(sce)[, c("sampleID.HTO", "genetic_donor")]))) %>%
  dplyr::select(1:length(donor_names)) %>% # discard columns of Doublets and Negatives
  slice_head(n=length(donor_names)) # discard rows of Doublets and Negatives

# match donor to sample ID according to the best match with HTO
sampleID.genetics <- setNames(c(rownames(tb)[apply(tb, MARGIN = 2, which.max)],
                                "Doublet","Negative"),
                              c(colnames(tb),
                                "Doublet","Negative"))
sampleID.genetics <- sampleID.genetics[sce$genetic_donor]
names(sampleID.genetics) <- colnames(sce)
sce$sampleID.genetics <- factor(sampleID.genetics, 
                                levels=c(samples,"Doublet","Negative"))

```


# 7 Visualize genetic demultiplexing results
## 7.1 Table of number of droplets

```{r}
janitor::tabyl(
  as.data.frame(colData(sce)[, c("HTODemux.ID", "genetic_donor")]),
  HTODemux.ID,
  genetic_donor) |>
  adorn_title(placement = "combined") |>
  adorn_totals("both") |>
  knitr::kable(
    caption = "Number of droplets assigned to each `HTO`/`Genetic donor` combination.")

```

## 7.2 Number of droplets (named by HTO and genetic donor)

```{r fig.width=10,fig.height=6}
# add colour
genetic_donor_colours <- setNames(
  c(palette.colors(nlevels(sce$genetic_donor)-2, "Set3"), "#d4a276","#9c6644"),
  levels(sce$genetic_donor))
sce$colours$genetic_donor_colours <- genetic_donor_colours[sce$genetic_donor]

# number of droplets assigned by HTO method
p1 <- ggcells(sce) + 
  geom_bar(aes(x = HTODemux.ID, fill = HTODemux.ID)) + 
  geom_text(stat='count', aes(x = HTODemux.ID, label=..count..), 
            hjust=1, size=2) +
  coord_flip() + 
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,2000),limits=c(0,14000)) +
  scale_fill_manual(values = sce$colours$hto_colours) 

p1.facet <- ggcells(sce) + 
  geom_bar(aes(x = HTODemux.ID, fill = HTODemux.ID)) + 
  geom_text(stat='count', aes(x = HTODemux.ID, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,8000,2000), limits = c(0,10000)) +
  scale_fill_manual(values = sce$colours$hto_colours) +
  theme(axis.title.y = element_blank()) +
  coord_flip()

# number of droplets assigned by genetic method
p2 <- ggcells(sce) + 
  geom_bar(aes(x = genetic_donor, fill = genetic_donor)) + 
  geom_text(stat='count', aes(x = genetic_donor, label=..count..,),
            hjust=1, size=2) +
  coord_flip() + 
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,2000),limits=c(0,14000)) +
  scale_fill_manual(values = sce$colours$genetic_donor_colours)

p2.facet <- ggcells(sce) + 
  geom_bar(aes(x = genetic_donor, fill = genetic_donor)) + 
  geom_text(stat='count', aes(x = genetic_donor, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,8000,2000), limits = c(0,10000)) +
  scale_fill_manual(values = sce$colours$genetic_donor_colours) +
  theme(axis.title.y = element_blank()) +
  coord_flip()

(p1 + p1.facet+plot_layout(width=c(1,2))) /
  (p2 + p2.facet+plot_layout(width=c(1,2))) +
  plot_layout(guides="collect")

```

## 7.3 Proportion of droplets (named by HTO and genetic donor)

```{r fig.width=10,fig.height=6}
# proportion of genetically assigned droplets in each HTO
p3 <- ggcells(sce) + 
  geom_bar(
    aes(x = HTODemux.ID, fill = genetic_donor),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$genetic_donor_colours)

# proportion of HTO assigned droplets in each genetic donor
p4 <- ggcells(sce) + 
  geom_bar(
    aes(x = genetic_donor, fill = HTODemux.ID),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$hto_colours)

(p3 + p3 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) /
  (p4 + p4 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) +
  plot_layout(guides = "collect")

```


## 7.4 Number of droplets (named by sample ID)

```{r fig.width=10,fig.height=6}
# add colour
sampleID.HTO_colours <- setNames(
  c(palette.colors(nlevels(sce$sampleID.HTO)-2, "Tableau 10"), "#ced4da","#6c757d"),
  levels(sce$sampleID.HTO))
sce$colours$sampleID.HTO_colours <- sampleID.HTO_colours[sce$sampleID.HTO]

sampleID.genetics_colours <- setNames(
  c(palette.colors(nlevels(sce$sampleID.genetics)-2, "Set3"), "#d4a276","#9c6644"),
  levels(sce$sampleID.genetics))
sce$colours$sampleID.genetics_colours <- sampleID.genetics_colours[sce$sampleID.genetics]

# number of droplets assigned by HTO method
p5 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By HTO") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,2000),limits=c(0,14000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID"))

p5.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,8000,2000), limits = c(0,10000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
    coord_flip()

# number of droplets assigned by genetic method
p6 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By genetics") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,2000), limits = c(0,14000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=FALSE)

p6.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,8000,2000), limits = c(0,10000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
  coord_flip()

(p5+p5.facet+plot_layout(width=c(1,2))) /
  (p6+p6.facet+plot_layout(width=c(1,2))) +
  plot_layout(guides="collect")
```

## 7.5 Proportion of droplets (named by sample ID)

```{r fig.width=10,fig.height=6}
# proportion of genetically assigned droplets in each HTO
p7 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.HTO, fill = sampleID.genetics),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours)

# proportion of HTO assigned droplets in each genetic donor
p8 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.genetics, fill = sampleID.HTO),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours)

((p7 + ggtitle("By HTO")) +
    p7 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) /
  ((p8 + ggtitle("By genetics")) + 
     p8 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) +
  plot_layout(guides = "collect")

```

# 8 Save object

```{r}
demux_dir <- here("data","SCEs","demux")
if(!dir.exists(demux_dir)) {
  dir.create(demux_dir, recursive = TRUE)
}

out <- paste0(demux_dir,'/',
              paste0(batch_name,".cellbender.demux.SCE.rds"))

if(!file.exists(out)) saveRDS(sce, out)

```
