---
title: 'Quality control for PBMC single captures'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(DropletUtils)
library(here)
library(ggplot2)
library(Seurat)
library(cowplot)
library(patchwork)
library(scater)
library(dplyr)
library(forcats)
library(janitor)
library(stringr)
library(AnnotationHub)
library(ensembldb)
library(msigdbr)
library(Homo.sapiens)
library(qs)
library(xgboost)
library(pROC)
library(scDblFinder)
library(BiocParallel)
library(tibble)
library(future.apply)
library(gridExtra)
})
source(here("code","scds.R"))
set.seed(1990)

```

# 2 Set names for captures

```{r}
# sampleID
samples <- c("M1C191B",
             "cf-p15","cf-p14","cf-p6","cf-p5",
             "mabsc-p11","mabsc-p9","mabsc-p4","mabsc-p3",
             "mabscpd-p13","mabscpd-p12","mabscpd-p1"
)

# batch_name
batch_name <- "Pcap"

# Specify capture name
capture_names <- setNames(samples, samples)

```

# 3 Read CellBender outputs
Ambient RNA and empty droplets were removed by CellBender (see code/cellbender.sh).
```{r}
captures <- setNames(
  here("data",
       "cellbender",
       capture_names,
       paste0(capture_names,".cellbender_filtered.h5")),
  capture_names)

# read objects
outLst <- here("data","SCEs","doublet",paste0(batch_name,".qc.sceLst.qs"))
if (!file.exists(outLst)) {
  sceLst <- lapply(capture_names, function(cap) {
    sce <- read10xCounts(samples = captures[cap], col.names = TRUE)
    stopifnot(!anyDuplicated(colnames(sce)))
    
    sce <- splitAltExps(
      sce,
      rowData(sce)$Type,
      "Gene Expression")
    
    # Tidy up colData
    sce$Capture <- factor(sce$Sample)
    sce$Sample <- NULL
    
    # Number of droplets after CellBender
    dim(sce)
    
    # add sampleID
    sce$sampleID <- factor(cap)
    
    # change DelayedMatrix to dgCMatrix
    assay(sce, "counts") <- as(assay(sce, "counts"), "dgCMatrix")
    
    sce
  })
}

```

# 4 Add metadata and sex check
```{r fig.height=9,fig.width=12}
outLst <- here("data","SCEs","doublet",paste0(batch_name,".qc.sceLst.qs"))
if (!file.exists(outLst)) {
  sceLst <- lapply(sceLst, function(sce) {
    # read metadata.csv
    metadata <- read.csv(here("data",
                              "sample_sheets",
                              paste0(levels(sce$Capture),".metadata.csv")))
    
    i <- match(sce$sampleID, metadata$sampleID)
    
    # add patient demographics
    colData(sce) <- cbind(
      colData(sce),
      metadata[i,c("Age","Sex","Condition","Bronchiectasis")]
    )
    
    sce
  })
  
  pt.list <- lapply(sceLst, function(sce) {
      # sex check by the expression of XIST, a female-specific gene. 
    # this detects certain types of sample-mix-ups.
    plotExpression(
      sce, 
      "XIST", 
      x = "sampleID",
      colour_by = "Sex", 
      exprs_values = "counts",
      swap_rownames = "Symbol") +
      scale_colour_manual(
        values = c(
          "Female" = "deeppink3", 
          "Male" = "deepskyblue"),
        name = "Sex") +
      theme_cowplot() + 
      theme(axis.text.x = element_text(angle = 0, vjust = 1, hjust = 1),
            axis.title.x = element_blank()) -> p
    return(p)
  })
  
  do.call("grid.arrange", c(pt.list, ncol=4))
}
```

# 5 Add feature-based annotation
```{r}
if (!file.exists(outLst)) {
  ah <- AnnotationHub(cache="/group/canc2/anson/.cache/R/AnnotationHub", ask=FALSE)
  EnsDb.Hsapiens.v98 <- query(ah, c("EnsDb", "Homo Sapiens", 98))[[1]]
    
  sceLst <- lapply(sceLst, function(sce) {
    # make feature names unique
    rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ID, rowData(sce)$Symbol)
    
    # prepare ensembl v98 database
    ensdb_columns <- setNames(c("GENEBIOTYPE", "SEQNAME"),
                              paste0("ENSEMBL.",
                                     c("GENEBIOTYPE", "SEQNAME")))
    stopifnot(all(ensdb_columns %in% columns(EnsDb.Hsapiens.v98)))
    
    ensdb_df <- DataFrame(
      lapply(ensdb_columns, function(column) {
        mapIds(
          x = EnsDb.Hsapiens.v98,
          keys = rowData(sce)$ID,
          keytype = "GENEID",
          column = column,
          multiVals = "first")
      }),
      row.names = rowData(sce)$ID)
    
    # prepare ncbi database
    ncbi_columns <- setNames(c("ALIAS", "ENTREZID", "GENENAME"),
                             paste0("NCBI.", c("ALIAS", "ENTREZID", "GENENAME")))
    stopifnot(all(ncbi_columns %in% columns(Homo.sapiens)))
    ncbi_df <- DataFrame(
      lapply(ncbi_columns, function(column) {
        mapIds(
          x = Homo.sapiens,
          keys = rowData(sce)$ID,
          keytype = "ENSEMBL",
          column = column,
          multiVals = "CharacterList")
      }),
      row.names = rowData(sce)$ID)
    
    rowData(sce) <- cbind(rowData(sce), ensdb_df, ncbi_df)
    
    sce
  })
}
```

# 6 QC 
## 6.1 Calculate metrics
```{r}
if (!file.exists(outLst)) {
  sceLst <- lapply(sceLst, function(sce) {
    # prepare gene sets
    ## mitochondrial gene set
    mito_set <- rownames(sce)[which(rowData(sce)$ENSEMBL.SEQNAME == "MT")]
    is_mito <- rownames(sce) %in% mito_set
    summary(is_mito)
    
    ## ribosomal gene set
    ribo_set <- grep("^RP(S|L)", rownames(sce), value = TRUE)
    c2_sets <- msigdbr(species = "Homo sapiens", category = "C2")
    ribo_set <- union(
      ribo_set,
      c2_sets[c2_sets$gs_name == "KEGG_RIBOSOME", ]$human_gene_symbol)
    is_ribo <- rownames(sce) %in% ribo_set
    summary(is_ribo)
    
    # calculate QC metrics
    sce <- addPerCellQCMetrics(
      sce,
      subsets = list(Mito = which(is_mito),
                     Ribo = which(is_ribo)), 
      use.altexps = NULL)
    
    sce
  })
  qsave(sceLst, file=outLst, nthreads=16)
} else {
  sceLst <- qread(outLst, nthreads=16)
}
```

## 6.2 Visualize metrics before removal
```{r fig.height=6,fig.width=6}
sample_colours <- setNames(
    c(palette.colors(10,"Tableau 10"),palette.colors(length(sceLst)-10,"Accent")),
    names(sceLst))

# add sample colours
sceLst <- lapply(sceLst, function(sce) {
  cols <- setNames(
    c(sample_colours[levels(sce$sampleID)]),levels(sce$sampleID))
  # add colour
  sce$colours <- S4Vectors::make_zero_col_DFrame(ncol(sce))
  sce$colours$sample_colours <- cols[sce$sampleID]
  sce
})

# library size
pt.lib <- lapply(sceLst, function(sce) {
  p1 <- plotColData(
    sce,
    "sum",
    x = "sampleID",
    colour_by = "sampleID",
    point_size = 0.5) +
    scale_y_log10() +
    scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
    theme(axis.title.x = element_blank()) + NoLegend()
  return(p1)
})

# number of genes detected
pt.ng <- lapply(sceLst, function(sce) {
  p2 <- plotColData(
    sce,
    "detected",
    x = "sampleID",
    colour_by = "sampleID",
    point_size = 0.5) +
    scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
    theme(axis.title.x = element_blank()) + NoLegend()
  return(p2)
})


do.call("grid.arrange", c(pt.lib, ncol=4))
do.call("grid.arrange", c(pt.ng, ncol=4))
```

## 6.3 Remove cells with features < 200
```{r}
sceLst_pre_QC_outlier_removal <- sceLst

sceLst_pre_QC_outlier_removal <- lapply(sceLst_pre_QC_outlier_removal, function(sce){
  feature_drop <- sce$detected < 200
  keep <- !feature_drop
  sce$keep <- keep
  sce
})

sceLst <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  sce <- sce[, sce$keep]
  sce
})

# print results
do.call("rbind",lapply(names(sceLst), function(s) {
  feature_drop <- sceLst_pre_QC_outlier_removal[[s]]$detected < 200
  data.frame(
  ByFeature = tapply(
    feature_drop, 
    sceLst_pre_QC_outlier_removal[[s]]$Capture, 
    sum,
    na.rm = TRUE),
  Remaining = as.vector(unname(table(sceLst[[s]]$Capture))),
  PercRemaining = round(
    100 * as.vector(unname(table(sceLst[[s]]$Capture))) /
      as.vector(
        unname(
          table(sceLst_pre_QC_outlier_removal[[s]]$Capture))), 1)) |>
  tibble::rownames_to_column("Capture") |>
  dplyr::arrange(dplyr::desc(PercRemaining)) 
})) %>% knitr::kable(caption = "Number of droplets remaining.")

```

## 6.4 Visualize metrics after removal
```{r fig.width=8,fig.height=4}
# library size
pt.lib <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  p1 <- plotColData(
    sce,
    "sum",
    x = "sampleID",
    colour_by = "keep",
    point_size = 0.5) +
    theme(axis.title.x = element_blank())
  return(p1)
})

# number of genes detected
pt.ng <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  p2 <- plotColData(
    sce,
    "detected",
    x = "sampleID",
    colour_by = "keep",
    point_size = 0.5) +
    theme(axis.title.x = element_blank())
  return(p2)
})

do.call("grid.arrange", c(pt.lib, ncol=4))
do.call("grid.arrange", c(pt.ng, ncol=4))

```

# 7 Doublet detection
## 7.1 Call doublets
```{r}
outLst <- here("data","SCEs","doublet",paste0(batch_name,".qc.rmDbl.sceLst.qs"))
if (!file.exists(outLst)) {
  out <- here("data","SCEs","doublet",paste0(batch_name,".doublets.SCE.qs"))
  if(!file.exists(out)){
    sceLst_pre_QC_outlier_removal <- lapply(sceLst, function(sce){
      # scds
      sce1 <- bcds(sce, retRes = TRUE, estNdbl = TRUE)
      sce1 <- cxds(sce1, retRes = TRUE, estNdbl = TRUE)
      sce1 <- cxds_bcds_hybrid(sce1, estNdbl = TRUE)
      # scDblFInder 
      set.seed(1990)
      bp <- MulticoreParam(16, RNGseed=1990)
      sce1 <- scDblFinder(sce1, dbr = ncol(sce1)/1000*0.008, BPPARAM=bp)
      sce1
    })
    
    lapply(sceLst_pre_QC_outlier_removal, function(s){
      colData(s) %>% 
        data.frame %>%
        rownames_to_column(var = "cell")
    }) %>% 
      bind_rows() %>% 
      qsave(file = out, nthreads=16)
    qsave(sceLst_pre_QC_outlier_removal, file=outLst, nthreads=16)
  }
} else {
  sceLst_pre_QC_outlier_removal <- qread(outLst, nthreads=16)
}
```

## 7.2 Remove doublets
```{r}
sceLst <- lapply(sceLst_pre_QC_outlier_removal, function(sce){
  sce <- sce[, !(sce$hybrid_call & sce$scDblFinder.class == "doublet")]
})

# n doublet
do.call("rbind",lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  table(sce$Capture, sce$hybrid_call & sce$scDblFinder.class == "doublet") 
})) -> tab

tab %>%
  knitr::kable(caption = "n doublet called by both scds and scDblFinder")

# % doublet
data.frame(pct=round(tab[,2]/(tab[,1]+tab[,2])*100,
                       digits=3)) %>% 
  knitr::kable(caption = "% doublet called by both scds and scDblFinder")
```


# 8 QC: mitochondrial percent
## 8.1 Visualize metrics before removal
```{r fig.width=8,fig.height=6}
# subsets_Mito_percent
pt.mito <- lapply(sceLst, function(sce) {
  p1 <- plotColData(
    sce,
    "subsets_Mito_percent",
    x = "sampleID",
    colour_by = "sampleID",
    point_size = 0.5) +
    scale_y_log10() +
    scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
    theme(axis.title.x = element_blank()) 
  return(p1)
})

# number of genes detected
pt.ribo <- lapply(sceLst, function(sce) {
  p2 <- plotColData(
    sce,
    "subsets_Ribo_percent",
    x = "sampleID",
    colour_by = "sampleID",
    point_size = 0.5) +
    scale_y_log10() +
    scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
    theme(axis.title.x = element_blank())
  return(p2)
})

suppressWarnings(do.call("grid.arrange", c(pt.mito, ncol=4)))
suppressWarnings(do.call("grid.arrange", c(pt.ribo, ncol=4)))

```

## 8.2 Remove cells with >10% mitochondrial reads
```{r}
sceLst_pre_QC_outlier_removal <- sceLst

sceLst_pre_QC_outlier_removal <- lapply(sceLst_pre_QC_outlier_removal, function(sce){
  mito_drop <- sce$subsets_Mito_percent > 10
  keep <- !mito_drop
  sce$keep <- keep
  sce
})

sceLst <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  sce <- sce[, sce$keep]
  sce
})

# print results
do.call("rbind",lapply(names(sceLst), function(s) {
  mito_drop <- sceLst_pre_QC_outlier_removal[[s]]$subsets_Mito_percent > 10
  data.frame(
  ByMito = tapply(
    mito_drop, 
    sceLst_pre_QC_outlier_removal[[s]]$Capture, 
    sum,
    na.rm = TRUE),
  Remaining = as.vector(unname(table(sceLst[[s]]$Capture))),
  PercRemaining = round(
    100 * as.vector(unname(table(sceLst[[s]]$Capture))) /
      as.vector(
        unname(
          table(sceLst_pre_QC_outlier_removal[[s]]$Capture))), 1)) |>
  tibble::rownames_to_column("Capture") |>
  dplyr::arrange(dplyr::desc(PercRemaining))
})) %>% knitr::kable(caption = "Number of droplets remaining.")


```

## 8.3 Visualize metrics after removal
```{r fig.width=8,fig.height=6}
# subsets_Mito_percent
pt.mito <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  p1 <- plotColData(
    sce,
    "subsets_Mito_percent",
    x = "sampleID",
    colour_by = "keep",
    point_size = 0.5) +
    scale_y_log10() +
    theme(axis.title.x = element_blank())
  return(p1)
})

# number of genes detected
pt.ribo <- lapply(sceLst_pre_QC_outlier_removal, function(sce) {
  p2 <- plotColData(
    sce,
    "subsets_Ribo_percent",
    x = "sampleID",
    colour_by = "keep",
    point_size = 0.5) +
    scale_y_log10() +
    theme(axis.title.x = element_blank())
  return(p2)
})

suppressWarnings(do.call("grid.arrange", c(pt.mito, ncol=4)))
suppressWarnings(do.call("grid.arrange", c(pt.ribo, ncol=4)))

```

# 9 Number of droplets retained after QC
```{r fig.width=8,fig.height=6}
pt.n <- lapply(sceLst, function(sce) {
  # number of droplets assigned by HTO method
  p1 <- ggcells(sce) + 
    geom_bar(aes(x = sampleID, fill = sampleID)) + 
    geom_text(stat='count', aes(x = sampleID, label=..count..), hjust=1, size=2) +
    coord_flip() + 
    ggtitle(levels(sce$Capture)) +
    ylab("Number of droplets") + 
    theme_cowplot(font_size = 8) + 
    scale_y_continuous(breaks=seq(0,16000,4000),limits=c(0,16000)) +
    scale_fill_manual(values = sce$colours$sample_colours) +
    theme(axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
    guides(fill=guide_legend(title="sampleID")) + NoLegend()
  return(p1)
})

suppressWarnings(do.call("grid.arrange", c(pt.n, ncol=3)))
```

# 10 Save object
```{r}
prep_dir <- here("data","SCEs","preprocessed")
if(!dir.exists(prep_dir)) {
  dir.create(prep_dir, recursive = TRUE)
}

out <- here("data","SCEs","preprocessed",paste0(batch_name,".preprocessed.sceLst.qs"))
if(!file.exists(out)) {
  qsave(sceLst, out, nthreads=16)
}

```
