---
title: 'Quality control for PBMC pool: PP1'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(DropletUtils)
library(here)
library(ggplot2)
library(Seurat)
library(cowplot)
library(patchwork)
library(scater)
library(dplyr)
library(forcats)
library(janitor)
library(stringr)
library(AnnotationHub)
library(ensembldb)
library(msigdbr)
library(Homo.sapiens)
library(qs)
library(xgboost)
library(pROC)
library(scDblFinder)
library(BiocParallel)
library(tibble)
library(future.apply)
library(gridExtra)
})
source(here("code","scds.R"))
set.seed(1990)

```

# 2 Set names for this pool

```{r}
# Specify batch name
batch_name <- "PP1"
num_of_captures <- 3

# Specify capture name
capture_names <- c(paste0(batch_name, "-c",1:num_of_captures))
capture_names <- setNames(capture_names, capture_names)

# Assign sample ID to HTO ID
# Manually list sample names matching HTO id (in alphabetical order)
samples <- c("M1C198B", #HTO3
             "M1C208A", #HTO6
             "M1C166(1)", #HTO7
             "M1C199B", #HTO8
             "M1C207B", #HTO10
             "M1C190C", #HTO12
             "M1N087", #HTO13
             "M1N092" #HTO14
)

```

# 3 Filter out vireo/HTO-derived doublets 
## 3.1 Filter
```{r}
# read demultiplexing outputs
sce <- readRDS(here("data",
                    "SCEs",
                    "demux",
                    paste0(batch_name,".cellbender.demux.SCE.rds")))

# number of droplets before removal
dim(sce)

# A. remove droplets classified as "doublets" by Vireo
sce <- sce[, which(sce$sampleID.genetics != "Doublet")]

# number of droplets after doublet removal
dim(sce)

# look for the following droplets
## genetically non-negatives with matched sampleID.genetics and sampleID.HTO
## or classified as Doublet or Negative by HTODemux
nneg_match <- sce$sampleID.genetics != "Negative" & 
  (sce$sampleID.HTO == sce$sampleID.genetics | 
     sce$sampleID.HTO %in% c("Doublet","Negative"))

## genetically negatives but with HTO classification
neg_match <- sce$sampleID.genetics == "Negative" & 
  sce$sampleID.HTO %in% samples

# B. keep droplets that match the demultiplexing criteria
sce <- sce[, nneg_match | neg_match]

# number of droplets after demultiplexing
dim(sce)

# C. remove droplets with posterior counts of zero called by CellBender
# see: https://github.com/broadinstitute/CellBender/issues/111
sce <- sce[, colSums(counts(sce)) != 0]

# number of droplets after removing droplets with zero counts
dim(sce)

# rename sampleID
sampleID <- setNames(factor(case_when(
  sce$sampleID.genetics == "Negative" ~ sce$sampleID.HTO,
  TRUE ~ sce$sampleID.genetics),levels = samples),
  colnames(sce))

sce$sampleID <- sampleID
```

## 3.2 Visualize results after filtering
```{r}
# number of droplets assigned by HTO method
p1 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By HTO") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,3000),limits=c(0,16000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID"))

p1.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), 
            hjust=1, size=2) +
  #ggtitle("By HTO") +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,6000,2000), limits = c(0,6000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
    coord_flip()

# number of droplets assigned by genetic method
p2 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By genetics") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,12000,3000), limits = c(0,16000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=FALSE)

p2.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,6000,2000), limits = c(0,6000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
  coord_flip()

(p1+p1.facet+plot_layout(width=c(1,2))) /
  (p2+p2.facet+plot_layout(width=c(1,2))) +
  plot_layout(guides="collect")

# proportion of genetically assigned droplets in each HTO
p3 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.HTO, fill = sampleID.genetics),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours)

# proportion of HTO assigned droplets in each genetic donor
p4 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.genetics, fill = sampleID.HTO),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours)

((p3 + ggtitle("By HTO")) +
    p3 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) /
  ((p4 + ggtitle("By genetics")) + 
     p4 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) +
  plot_layout(guides = "collect")
```


# 4 Add metadata and sex check
```{r}
# read metadata.csv
metadata <- read.csv(here("data",
                          "sample_sheets",
                          paste0(batch_name,".metadata.csv")))

i <- match(sce$sampleID, metadata$sampleID)

# add patient demographics
colData(sce) <- cbind(
  colData(sce),
  metadata[i,c("Age","Sex","Condition","Bronchiectasis")]
)

# sex check by the expression of XIST, a female-specific gene. 
# this detects certain types of sample-mix-ups.
plotExpression(
  sce, 
  "XIST", 
  x = "sampleID",
  colour_by = "Sex", 
  exprs_values = "counts",
  swap_rownames = "Symbol") +
  scale_colour_manual(
    values = c(
      "Female" = "deeppink3", 
      "Male" = "deepskyblue"),
    name = "Sex") +
  theme_cowplot() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

```

# 5 Add feature-based annotation
```{r}
# make feature names unique
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ID, rowData(sce)$Symbol)

# prepare ensembl v98 database
ah <- AnnotationHub(cache="/group/canc2/anson/.cache/R/AnnotationHub", ask=FALSE)
EnsDb.Hsapiens.v98 <- query(ah, c("EnsDb", "Homo Sapiens", 98))[[1]]

ensdb_columns <- setNames(c("GENEBIOTYPE", "SEQNAME"),
                          paste0("ENSEMBL.",
                                 c("GENEBIOTYPE", "SEQNAME")))
stopifnot(all(ensdb_columns %in% columns(EnsDb.Hsapiens.v98)))

ensdb_df <- DataFrame(
  lapply(ensdb_columns, function(column) {
    mapIds(
      x = EnsDb.Hsapiens.v98,
      keys = rowData(sce)$ID,
      keytype = "GENEID",
      column = column,
      multiVals = "first")
  }),
  row.names = rowData(sce)$ID)

# prepare ncbi database
ncbi_columns <- setNames(c("ALIAS", "ENTREZID", "GENENAME"),
                         paste0("NCBI.", c("ALIAS", "ENTREZID", "GENENAME")))
stopifnot(all(ncbi_columns %in% columns(Homo.sapiens)))
ncbi_df <- DataFrame(
  lapply(ncbi_columns, function(column) {
    mapIds(
      x = Homo.sapiens,
      keys = rowData(sce)$ID,
      keytype = "ENSEMBL",
      column = column,
      multiVals = "CharacterList")
  }),
  row.names = rowData(sce)$ID)

rowData(sce) <- cbind(rowData(sce), ensdb_df, ncbi_df)

```

# 6 QC: nCount and nFeature
## 6.1 Calculate metrics
```{r}
# prepare gene sets
## mitochondrial gene set
mito_set <- rownames(sce)[which(rowData(sce)$ENSEMBL.SEQNAME == "MT")]
is_mito <- rownames(sce) %in% mito_set
summary(is_mito)

## ribosomal gene set
ribo_set <- grep("^RP(S|L)", rownames(sce), value = TRUE)
c2_sets <- msigdbr(species = "Homo sapiens", category = "C2")
ribo_set <- union(
  ribo_set,
  c2_sets[c2_sets$gs_name == "KEGG_RIBOSOME", ]$human_gene_symbol)
is_ribo <- rownames(sce) %in% ribo_set
summary(is_ribo)

# calculate QC metrics
sce <- addPerCellQCMetrics(
  sce,
  subsets = list(Mito = which(is_mito),
                 Ribo = which(is_ribo)), 
  use.altexps = NULL)

```

## 6.2 Visualize metrics before removal
```{r fig.width=10,fig.height=5}
# library size
p1 <- plotColData(
  sce,
  "sum",
  x = "sampleID",
  colour_by = "sampleID",
  point_size = 0.5) +
  scale_y_log10() +
  scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
  theme(axis.text.x = element_blank())

# number of genes detected
p2 <- plotColData(
  sce,
  "detected",
  x = "sampleID",
  colour_by = "sampleID",
  point_size = 0.5) +
  scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
  theme(axis.text.x = element_blank())

((p1 + NoLegend()) + p1 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names))))/
((p2 + NoLegend()) + p2 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names)))) + 
  plot_layout(guides="collect")

```

## 6.3 Remove cells with features < 200
```{r}
sce$batch <- interaction(
  sce$Capture, 
  sce$sampleID, 
  drop = TRUE,
  lex.order = FALSE)

feature_drop <- sce$detected < 200

sce_pre_QC_outlier_removal <- sce
keep <- !feature_drop
sce_pre_QC_outlier_removal$keep <- keep
sce <- sce[, keep]

data.frame(
  ByFeature = tapply(
    feature_drop, 
    sce_pre_QC_outlier_removal$batch, 
    sum,
    na.rm = TRUE),
  Remaining = as.vector(unname(table(sce$batch))),
  PercRemaining = round(
    100 * as.vector(unname(table(sce$batch))) /
      as.vector(
        unname(
          table(sce_pre_QC_outlier_removal$batch))), 1)) |>
  tibble::rownames_to_column("batch") |>
  dplyr::arrange(dplyr::desc(PercRemaining)) |>
  DT::datatable(
    caption = "Number of droplets removed by each QC step and the number of droplets remaining.",
    rownames = FALSE) |>
  DT::formatRound("PercRemaining", 1)

```

## 6.4 Visualize metrics after removal
```{r fig.width=10,fig.height=5}
p3 <- plotColData(
  sce_pre_QC_outlier_removal,
  "sum",
  x = "sampleID",
  colour_by = "keep",
  point_size = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p4 <- plotColData(
  sce_pre_QC_outlier_removal,
  "detected",
  x = "sampleID",
  colour_by = "keep",
  point_size = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

((p3 + NoLegend()) + p3 + facet_grid(~sce_pre_QC_outlier_removal$Capture) + plot_layout(widths=c(1, length(capture_names))))/
((p4 + NoLegend()) + p4 + facet_grid(~sce_pre_QC_outlier_removal$Capture) + plot_layout(widths=c(1, length(capture_names)))) + 
  plot_layout(guides="collect")

```

# 7 Doublet detection
## 7.1 Call doublets
```{r}
outLst <- here("data","SCEs","doublet",paste0(batch_name,".rmDbl.sceLst.qs"))

if (!file.exists(outLst)) {
  out <- here("data","SCEs","doublet",paste0(batch_name,".doublets.SCE.qs"))
  if(!file.exists(out)){
    sceLst <- lapply(capture_names, function(cn){
      sce1 <- sce[, sce$Capture == cn]
      # scds
      sce1 <- bcds(sce1, retRes = TRUE, estNdbl = TRUE)
      sce1 <- cxds(sce1, retRes = TRUE, estNdbl = TRUE)
      sce1 <- cxds_bcds_hybrid(sce1, estNdbl = TRUE)
      # scDblFInder 
      set.seed(1990)
      bp <- MulticoreParam(16, RNGseed=1990)
      sce1 <- scDblFinder(sce1, dbr = ncol(sce1)/1000*0.008, BPPARAM=bp)
      sce1
    })
    
    lapply(sceLst, function(s){
      colData(s) %>% 
        data.frame %>%
        rownames_to_column(var = "cell")
    }) %>% 
      bind_rows() %>% 
      qsave(file = out, nthreads=16)
    qsave(sceLst, file=outLst, nthreads=16)
  }
} else {
  sceLst <- qread(outLst, nthreads=16)
}

# add per-capture doublet stat to sce
columns <- c("bcds_score","bcds_call","cxds_score","cxds_call",
             "hybrid_score","hybrid_call","scDblFinder.class","scDblFinder.score",
             "scDblFinder.weighted","scDblFinder.cxds_score")

do.call("rbind",lapply(sceLst, function(sce) {
  colData(sce)[,columns]
})) -> doublet_df

colData(sce) <- cbind(
  colData(sce),
  doublet_df
)
```

## 7.2 Remove doublets
```{r}
doublet_drop <- sce$hybrid_call & sce$scDblFinder.class == "doublet"

sce_pre_QC_outlier_removal <- sce
keep <- !doublet_drop
sce_pre_QC_outlier_removal$keep <- keep

sce <- sce[, keep]

data.frame(
  ByDoublet = tapply(
    doublet_drop, 
    sce_pre_QC_outlier_removal$batch, 
    sum,
    na.rm = TRUE),
  Remaining = as.vector(unname(table(sce$batch))),
  PercRemaining = round(
    100 * as.vector(unname(table(sce$batch))) /
      as.vector(
        unname(
          table(sce_pre_QC_outlier_removal$batch))), 1)) |>
  tibble::rownames_to_column("batch") |>
  dplyr::arrange(dplyr::desc(PercRemaining)) |>
  DT::datatable(
    caption = "Number of droplets removed by each QC step and the number of droplets remaining.",
    rownames = FALSE) |>
  DT::formatRound("PercRemaining", 1)

```



# 8 QC: mitochondrial percent
## 8.1 Visualize metrics before removal
```{r}
# subsets_Mito_percent
p1 <- plotColData(
  sce,
  "subsets_Mito_percent",
  x = "sampleID",
  colour_by = "sampleID",
  point_size = 0.5) +
  scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
  theme(axis.text.x = element_blank())

# subsets_Ribo_percent
p2 <- plotColData(
  sce,
  "subsets_Ribo_percent",
  x = "sampleID",
  colour_by = "sampleID",
  point_size = 0.5) +
  scale_colour_manual(values = sce$colours$sample_colours, name = "sampleID") +
  theme(axis.text.x = element_blank())

((p1 + NoLegend()) + p1 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names))))/
((p2 + NoLegend()) + p2 + facet_grid(~sce$Capture) + plot_layout(widths=c(1, length(capture_names)))) + 
  plot_layout(guides="collect")

```

## 8.2 Remove cells with >10% mitochondrial reads
```{r}
mito_drop <- sce$subsets_Mito_percent > 10

sce_pre_QC_outlier_removal <- sce
keep <- !mito_drop
sce_pre_QC_outlier_removal$keep <- keep
sce <- sce[, keep]

data.frame(
  ByMito = tapply(
    mito_drop, 
    sce_pre_QC_outlier_removal$batch, 
    sum,
    na.rm = TRUE),
  Remaining = as.vector(unname(table(sce$batch))),
  PercRemaining = round(
    100 * as.vector(unname(table(sce$batch))) /
      as.vector(
        unname(
          table(sce_pre_QC_outlier_removal$batch))), 1)) |>
  tibble::rownames_to_column("batch") |>
  dplyr::arrange(dplyr::desc(PercRemaining)) |>
  DT::datatable(
    caption = "Number of droplets removed by each QC step and the number of droplets remaining.",
    rownames = FALSE) |>
  DT::formatRound("PercRemaining", 1)

```

## 8.3 Visualize metrics after removal
```{r fig.width=10,fig.height=5}
p3 <- plotColData(
  sce_pre_QC_outlier_removal,
  "subsets_Mito_percent",
  x = "sampleID",
  colour_by = "keep",
  point_size = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4 <- plotColData(
  sce_pre_QC_outlier_removal,
  "subsets_Ribo_percent",
  x = "sampleID",
  colour_by = "keep",
  point_size = 0.5) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

((p3 + NoLegend()) + p3 + facet_grid(~sce_pre_QC_outlier_removal$Capture) + plot_layout(widths=c(1, length(capture_names))))/
((p4 + NoLegend()) + p4 + facet_grid(~sce_pre_QC_outlier_removal$Capture) + plot_layout(widths=c(1, length(capture_names)))) + 
  plot_layout(guides="collect")

```

# 9 Droplets retained 
## 9.1 Number (named by sample ID)
```{r fig.width=10,fig.height=6}
# number of droplets assigned by HTO method
p1 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By HTO") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,16000,4000),limits=c(0,16000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID"))

p1.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.HTO, fill = sampleID.HTO)) + 
  geom_text(stat='count', aes(x = sampleID.HTO, label=..count..), 
            hjust=1, size=2) +
  #ggtitle("By HTO") +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,4000,1000), limits = c(0,4000)) +
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
    coord_flip()

# number of droplets assigned by genetic method
p2 <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), hjust=1, size=2) +
  coord_flip() + 
  ggtitle("By genetics") +
  ylab("Number of droplets") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,16000,4000), limits = c(0,16000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=FALSE)

p2.facet <- ggcells(sce) + 
  geom_bar(aes(x = sampleID.genetics, fill = sampleID.genetics)) + 
  geom_text(stat='count', aes(x = sampleID.genetics, label=..count..), 
            hjust=1, size=2) +
  ylab("Number of droplets") + 
  facet_grid(~Capture, scales = "fixed", space = "fixed") + 
  theme_cowplot(font_size = 8) + 
  scale_y_continuous(breaks=seq(0,4000,1000), limits = c(0,4000)) +
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours) +
  theme(axis.title.y = element_blank()) +
  guides(fill=guide_legend(title="sampleID")) +
  coord_flip()

(p1+p1.facet+plot_layout(width=c(1,2))) /
  (p2+p2.facet+plot_layout(width=c(1,2))) +
  plot_layout(guides="collect")

```

## 9.2 Proportion (named by sample ID)
```{r fig.width=10,fig.height=6}
# proportion of genetically assigned droplets in each HTO
p3 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.HTO, fill = sampleID.genetics),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.genetics_colours)

# proportion of HTO assigned droplets in each genetic donor
p4 <- ggcells(sce) + 
  geom_bar(
    aes(x = sampleID.genetics, fill = sampleID.HTO),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  theme_cowplot(font_size = 8) + 
  scale_fill_manual(values = sce$colours$sampleID.HTO_colours)

((p3 + ggtitle("By HTO")) +
    p3 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) /
  ((p4 + ggtitle("By genetics")) + 
     p4 + facet_grid(~Capture) + plot_layout(widths = c(1, 2))) +
  plot_layout(guides = "collect")

```

# 10 Save object
```{r}
prep_dir <- here("data","SCEs","preprocessed")
if(!dir.exists(prep_dir)) {
  dir.create(prep_dir, recursive = TRUE)
}

out <- paste0(prep_dir,'/',
              paste0(batch_name,".preprocessed.SCE.qs"))

if(!file.exists(out)) qsave(sce, out, nthreads=16)

```
