---
title: 'PBMC manual annotation'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(speckle)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(rstatix)
library(ggpubr)
library(rcartocolor)
library(readr)
})
set.seed(1990)

```

# 2 Reference-based annotation
Reference-based annotation was performed on the clustered dataset using three references:
a. Yoshida et al. 2021. For details, refer to analysis/04a.annotation_Yoshida.Rmd

b. Stephenson et al. 2021. For details, refer to analysis/04b.annotation_Stephenson.Rmd

c. Azimuth PBMC using RunAzimuth(). Due to internal in-house server compatibility issues, R v4.2.2 instead of R v4.1.2 was used for running this step. For details, refer to code/azimuth_pbmc.R

## 2.1 Load object
```{r}
out <- here("data","SCEs","annotation",
            "pbmc.preprocessed.merged.integrated.clustered.Yoshida.Stephenson.Azimuth.rd.SEU.qs")
if (!file.exists(out)) {
  seu <- qread(here("data",
                    "SCEs",
                    "annotation",
                    "pbmc.preprocessed.merged.integrated.clustered.Yoshida.Stephenson.Azimuth.SEU.qs"),
               nthreads=16)
  DefaultAssay(seu) <- "SCT"
  # remove some info to reduce the file size
  seu@assays$prediction.score.annotation_broad <- NULL
  seu@assays$prediction.score.annotation_detailed <- NULL
  seu@assays$prediction.score.celltype.l1 <- NULL
  seu@assays$prediction.score.celltype.l2 <- NULL
  seu@assays$prediction.score.celltype.l3 <- NULL
  seu@assays$prediction.score.initial_clustering <- NULL
  seu@assays$prediction.score.author_cell_type <- NULL
  seu@assays$refAssay <- NULL
  gc()
  
  # save object with reduced file size
  qsave(seu, file=out, nthreads=16)
  
} else {
  seu <- qread(out,nthreads=16)
}
seu.backup <- seu
```

## 2.2 Prepare object for downstream analysis
```{r}
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("eAIR003","eAIR005","eAIR027","M1N087","M1N092", # healthy
                                "M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B", # CFN
                                "M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A" # CFB
                                ))

# Add Condition
Condition <- factor(
      case_when(
        seu$sampleID %in% c("eAIR003","eAIR005","eAIR027","M1N087","M1N092") ~ "C",
        seu$sampleID %in% c("M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B",
                            "M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A") ~ "CF"))

names(Condition) <- colnames(seu)
seu <- AddMetaData(seu, Condition, col.name = "Condition")
seu$Condition <- factor(seu$Condition,levels=c("C","CF"))

# Add Condition2
Condition2 <- factor(
      case_when(
        seu$sampleID %in% c("eAIR003","eAIR005","eAIR027","M1N087","M1N092") ~ "C",
        seu$sampleID %in% c("M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B") ~ "CFN",
        seu$sampleID %in% c("M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A") ~ "CFB"))

names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFN","CFB"))

# Set ID for this study
ID <- setNames(c(paste0("C-",1:5),paste0("CF-",1:12)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],levels=c(paste0("C-",1:5),paste0("CF-",1:12)
                 )),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set ID2 for this study
ID2 <- setNames(c(paste0("C-",1:5),paste0("CFN-",1:6),paste0("CFB-",1:6)),
               levels(seu$sampleID))

ID2 <- setNames(factor(ID2[seu$sampleID],
                       levels=c(paste0("C-",1:5),paste0("CFN-",1:6),paste0("CFB-",1:6))),
               colnames(seu))

seu <- AddMetaData(seu, ID2, col.name = "ID2")

# Re-order levels of SCT_snn_res.3
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.3, levels=as.character(seq(1, length(levels(seu$SCT_snn_res.3)))))

# Number of droplets after annotation
dim(seu)

# Number of droplets per sample
summary(seu$ID2)
```

# 3. Visualize annotation
```{r}
# prepare colours
colours <- c(#carto_pal(12,"Vivid"), carto_pal(12,"Safe"),
          carto_pal(12,"Pastel")[2:4],
          carto_pal(12,"Pastel")[1],
          carto_pal(12,"Pastel")[5:11],
          carto_pal(12,"Prism")[1:11], carto_pal(12,"Safe")[1:11],
          carto_pal(12,"Bold")[1:11])

Condition3.colours <- setNames(c("#26547c","#ef476f","#ffd166"),
                               c("C","CFB","CFN"))
```

## 3.1 UMAP 
### 3.1.1 Not split
```{r fig.width=8,fig.height=12}

group <- setNames(c("Yoshida_broad","Yoshida_detailed",
                    "Stephenson_detailed",
                    "Azimuth_l2","Azimuth_l3",
                    "SCT_snn_res.3"),
                  c("predicted.annotation_broad","predicted.annotation_detailed",
                    "predicted.author_cell_type","predicted.celltype.l2",
                    "predicted.celltype.l3",
                    "SCT_snn_res.3")
                  )
plots <- list()
plots <- sapply(names(group), FUN = function(x) {
  p <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=rep(colours,2),
        group.by = x) + NoLegend() +
  labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
  return(p)
})

((plots[[1]] + plots[[2]] + plot_layout(widths=c(1, 1)))/
    (plots[[3]] + plots[[4]] + plot_layout(widths=c(1, 1)))/
    (plots[[5]] + plots[[6]] + plot_layout(widths=c(1, 1)))) +
  plot_layout(guides="collect")

```

### 3.1.2 Split
```{r}
group <- setNames(c("Yoshida_broad","Yoshida_detailed",
                    "Stephenson_broad","Stephenson_detailed",
                    "Azimuth_l2",
                    "SCT_snn_res.3"),
                  c("predicted.annotation_broad","predicted.annotation_detailed",
                    "predicted.initial_clustering","predicted.author_cell_type",
                    "predicted.celltype.l2",
                    "SCT_snn_res.3")
                  )
plots <- list()
plots <- sapply(names(group), FUN = function(x) {
  p <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        split.by = "Condition3",
        group.by = x) + #NoLegend() +
  labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
  return(p)
})
```

::::: {.panelset}

::: {.panel}

#### Yoshida {.panel-name}

```{r fig.width=12,fig.height=8}
(plots[[1]]/plots[[2]]) + plot_layout(guides="collect")

```

:::

::: {.panel}

#### Stephenson {.panel-name}

```{r fig.width=12,fig.height=8}
(plots[[3]]/plots[[4]]) + plot_layout(guides="collect")

```

:::

::: {.panel}

#### Azimuth {.panel-name}

```{r fig.width=12,fig.height=4}
plots[[5]] + plot_layout(guides="collect")

```

:::

::: {.panel}

#### SCT_snn_res.3 {.panel-name}

```{r fig.width=12,fig.height=4}
plots[[6]] + plot_layout(guides="collect")

```

:::

:::::

#### Show legend 
```{r fig.height=5, fig.width=6, fig.dpi=800}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        group.by = "SCT_snn_res.3") + #NoLegend() +
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
```

### 3.1.3 Check batch effect
::::: {.panelset}

::: {.panel}

#### By batch {.panel-name}
```{r fig.width=12,fig.height=12}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, ncol = 5,
        cols=colours,
        split.by = "ID2",
        group.by = "SCT_snn_res.3") + #NoLegend() +
  #labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
```
:::

::: {.panel}

#### By Condition {.panel-name}
```{r fig.width=9,fig.height=4}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, ncol = 5,
        cols=colours,
        split.by = "Condition3",
        group.by = "SCT_snn_res.3") + #NoLegend() +
  #labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
```

:::
:::::

## 3.2 Proportion of cells per cluster
```{r}
labels <- group
p <- vector("list",length(labels))

for(label in names(labels)){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count.., vjust=0.5, angle=45), size=3, stat = "count",
              colour = "black") +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = colours) +
    labs(y = "# cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition3) %>%
    group_by(!!sym(label), Condition3) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition3)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition3") + scale_fill_manual(values = Condition3.colours)-> p2
  
  p1 / p2 -> p[[label]]
}

```

::::: {.panelset}

::: {.panel}

### Yoshida_broad {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.annotation_broad + plot_layout(guides="collect")

```

:::

::: {.panel}

### Yoshida_detailed {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.annotation_detailed + plot_layout(guides="collect")

```

:::

::: {.panel}

### Stephenson_broad {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.initial_clustering + plot_layout(guides="collect")

```

:::

::: {.panel}

### Stephenson_detailed {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.author_cell_type + plot_layout(guides="collect")

```

:::

::: {.panel}

### Azimuth {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.celltype.l2 + plot_layout(guides="collect")

```

:::

::: {.panel}

### SCT_snn_res.3 {.panel-name}

```{r fig.width=8,fig.height=6}
p$SCT_snn_res.3 + plot_layout(guides="collect")

```

:::

:::::

## 3.3 Marker gene plots
```{r}
# factorize metadata
seu$predicted.annotation_broad <- factor(seu$predicted.annotation_broad)
seu$predicted.annotation_detailed <- factor(seu$predicted.annotation_detailed)
seu$predicted.initial_clustering <- factor(seu$predicted.initial_clustering)
seu$predicted.author_cell_type <- factor(seu$predicted.author_cell_type)
seu$predicted.celltype.l2 <- factor(seu$predicted.celltype.l2)

# Prepare marker csv
celltype.list <- list(yoshida_broad = levels(seu$predicted.annotation_broad),
                 yoshida_detailed = levels(seu$predicted.annotation_detailed),
                 stephenson_broad = levels(seu$predicted.initial_clustering),
                 stephenson_detailed = levels(seu$predicted.author_cell_type),
                 azimuth = levels(seu$predicted.celltype.l2))

df.list <- lapply(names(celltype.list), FUN = function(x) {
  d <- data.frame(reference = rep(names(celltype.list[x]),length(celltype.list[[x]])),
                  celltype = celltype.list[[x]])
  return(d)
})

df <- bind_rows(df.list)

if (!dir.exists(here("data","annotation"))) {dir.create(here("data","annotation"))}
out <- here("data","annotation","RefCelltype.csv")
if (!file.exists(out)) {write.csv(df, file = out, row.names = FALSE)}

# Gene list
## Stephenson 2021, derived from FigS6E
stephenson.genes <- c("CD3D","IL7R","CCR7","LTB","LEF1","SELL","TIGIT","FOXP3","CD4","CD8A","CD8B","KLRB1","ITGB1",
                      "CCL5","TRGV9","TRDV2","TRAV1-2","GNLY",
                      "GZMB","FCGR3A","NCAM1","MS4A1","CD79A","IGHM","IGHG1","IGKC","CD38","CD14",
                      "S100A8","FCER1G","CST3","MS4A7","HLA-DQA1","HLA-DQB1","CD1C","LILRA4","IL3RA",
                      "IRF7","CD34","SPINK2","HBA2","HBB","PF4","PPBP","TOP2A","MKI67","IFI44","IFIT3","IFI6")

## Yoshida 2021, derived from FigS6
yoshida.genes <- c("CD3D","CD4","CD8A","CCR7","CD27","SELL","CX3CR1","IL7R","GZMH","PRF1",
                   "TRGV9","TRDV2","FOXP3","IL2RA","TRAV7","TRAV1-2","SLC4A10","NCR1","NCAM1","GNLY",
                   "TNFRSF18","TNFRSF4","FCER1G","CD14","FCGR3A","IL6","CLEC4C","IL3RA","AXL","SIGLEC6",
                   "CLEC9A","FCER1A","FCER2","IGHD","CD19","CD24","TCL1A","IGHM","CD79A","MS4A1",
                   "TNFRSF13B","CR2","BANK1","JCHAIN","IGHG1","TBX21","FCRL5","FCRL3","ENTPD1","KIT",
                   "CD34","SPINK2","TPSAB1","TPSB2","PRG2","EPX","MKI67","PPBP","PF4","HBB")

```

### 3.3.1 DotPlot
::::: {.panelset}

::: {.panel}

#### Stephenson {.panel-name}

```{r fig.width=18,fig.height=15}
p2 <- DotPlot(seu,
        features = stephenson.genes,
        group.by = "SCT_snn_res.3",
        col.max = 2,
        col.min = 0,
        dot.scale = 5
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Ariel"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.bugn(9)[1], high=brewer.bugn(9)[8]) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p2$guides$colour$title <- "Mean Expression"
p2$guides$colour$title.position <- "top"
p2$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p2))

```

:::

::: {.panel}

#### Yoshida {.panel-name}

```{r fig.width=21,fig.height=15}
p2 <- DotPlot(seu,
        features = yoshida.genes,
        group.by = "SCT_snn_res.3",
        col.max = 2,
        col.min = 0,
        dot.scale = 5
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Arial"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.blues(9)[1], high=brewer.blues(9)[8]) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p2$guides$colour$title <- "Mean Expression"
p2$guides$colour$title.position <- "top"
p2$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p2))

```

:::

::: {.panel}

#### T cell {.panel-name}

```{r fig.width=5,fig.height=12}
s <- subset(seu, SCT_snn_res.3 %in% c("10","11","12","20","21","37","38","52","54","53","50",
                                      "41","47","49","32","3","18","43"),invert=TRUE)
s$SCT_snn_res.3 <- fct_drop(s$SCT_snn_res.3)

genes <- c("CD4","CD40LG","CD8A","CD8B")
p2 <- DotPlot(s,
        features = genes,
        group.by = "SCT_snn_res.3",
        col.max = 2,
        col.min = 0,
        dot.scale = 5
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Ariel"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.bugn(9)[1], high=brewer.bugn(9)[8]) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p2$guides$colour$title <- "Mean Expression"
p2$guides$colour$title.position <- "top"
p2$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p2))

```

:::

:::::


### 3.3.2 FeaturePlot
```{r fig.width=9,fig.height=6}
# FeaturePlot
FeaturePlot(seu, c("CD3D","TRAC","CD4", "CD8A","TRDC","TRAV1-2"),
            ncol = 3, raster=TRUE, cols=brewer.reds(9)[2:9]) & NoLegend()

```

## 3.4 Cepo
```{r}
# Run Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$SCT_snn_res.3,
                    exprsPct = 0.2)
cepoMarkers$stats

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:100]
}) %>% data.frame() -> dat

colnames(dat) <- gsub("^X","",colnames(cepoMarkers$stats))

# Get the column names of your data frame
current_colnames <- colnames(dat)

# Convert the names to numeric type, sort them, and then convert back to character
sorted_colnames <- as.character(sort(as.numeric(current_colnames)))

# Reorder the 'dat' data frame using the numerically sorted column names
dat <- dat[, sorted_colnames]

# Verify the new order
head(dat)

if (!dir.exists(here("data","Cepo"))) {
  dir.create(here("data","Cepo"))
}
dat %>% data.frame(check.names = FALSE) %>% write.csv(file = here("data","Cepo","pbmc.markers.SCT_snn_res.3.csv"),row.names = FALSE)

```

# 4. Manual annotation
## 4.1 Annotation
### 4.1.1 Level 1
```{r}
# Use Azimuth result to annotate mast cells
draft1 <- factor(
      case_when(
        seu$predicted.celltype.l2 == "HSPC" ~ "HSPC",
        seu$SCT_snn_res.3 == "50" & seu$predicted.celltype.l2 == "pDC" ~ "pDC",
        seu$SCT_snn_res.3 == "33" & seu$predicted.celltype.l2 == "NK_CD56bright" ~ "NK.CD56hi"))

seu <- AddMetaData(seu, draft1, col.name = "draft1")

# SCT_snn_res.3 was used
draft2 <- setNames(c("CD4T","CD4T","NK.CD56lo","CD4T","CD8T", # 1-5
                     "CD4T","CD4T","CD8T","CD8T","B", # 6-10
                     "B","B","CD4T","CD4T","CD4T", # 11-15
                     "CD4T","CD8T","NK.CD56lo","CD4T","B", # 16-20
                     "B","CD8T","CD4T","CD4T","CD8T", # 21-25
                     "CD8T","CD4T","CD8T","CD4T","CD4T", # 26-30
                     "CD4T","NK.CD56lo","NK.CD56lo","CD4T","CD8T", # 31-35
                     "CD8T","B","B","CD4T","CD4T", # 36-40
                     "CD14.mono","CD4T","NK.CD56lo","CD8T","CD8T", # 41-45
                     "CD8T","CD16.mono","CD4T","NK.CD56lo","cDC", # 46-50
                     "CD8T","Plasmablast","ILC","Prolif.lymph" # 51-54
                     ), levels(seu$SCT_snn_res.3))

draft2 <- factor(setNames(draft2[seu$SCT_snn_res.3],colnames(seu)))

seu <- AddMetaData(seu, draft2, col.name = "draft2")

# coalesce two draft columns
seu@meta.data %>%
  mutate(ann_lv1=coalesce(draft1,draft2)) -> seu@meta.data

# reorder levels
seu$ann_lv1 <- factor(seu$ann_lv1,
                       levels=c("CD4T","CD8T","NK.CD56hi","NK.CD56lo","Prolif.lymph","ILC",
                                "CD14.mono", "CD16.mono","cDC","pDC","B",
                                "Plasmablast","HSPC"))

# remove draft columns
seu$draft1 <- NULL
seu$draft2 <- NULL

# Number of cells per celltype
summary(seu$ann_lv1)

seu.backup <- seu
```
### 4.1.2 Level 2
```{r}
# PBMC cell names (must be length 110809)
cells <- rownames(seu@meta.data)   # or colnames(seu)

# T cell map from csv: cell -> subtype (KEEP Ambiguous)
ann <- read_csv(
  here("data","annotation","tcell_subtype_annotations.csv"),
  show_col_types = FALSE
) |>
  distinct(cell, .keep_all = TRUE) |>
  mutate(subtype = as.character(subtype))   # no na_if()

t_map <- setNames(ann$subtype, ann$cell)

# B cell cluster ids (same length as PBMC cells)
b_clust <- as.integer(as.character(seu$SCT_snn_res.3))

# draft1: baseline ann_lv1 for everyone
draft1 <- as.character(seu$ann_lv1)

# draft2: overwrite labels for T and B cells; else NA
draft2 <- case_when(
  cells %in% names(t_map) ~ unname(t_map[cells]),  # will return "Ambiguous" if that's in the csv
  seu$ann_lv1 == "B" & b_clust %in% c(10, 11, 12, 37, 38) ~ "B.naive",
  seu$ann_lv1 == "B" & b_clust == 21                      ~ "B.intermediate",
  seu$ann_lv1 == "B" & b_clust == 20                      ~ "B.memory",
  TRUE ~ NA_character_
)

# subtype where available; otherwise ann_lv1
seu$ann_lv2 <- coalesce(draft2, draft1)

# Factorize (include Ambiguous level)
seu$ann_lv2 <- factor(
  seu$ann_lv2,
  levels = c(
    "CD4T.naive","CD4T.CM","CD8T.naive","CD8T.naive/CM",
    "CD4T.EM","CD8T.EM","Treg",
    "gdT","MAIT","dnT",
    "NKT","NK.CD56hi","NK.CD56lo",
    "Prolif.lymph","ILC",
    "B.naive","B.intermediate","B.memory","Plasmablast",
    "CD14.mono","CD16.mono","cDC","pDC","HSPC",
    "Ambiguous"
  )
)

summary(seu$ann_lv2)
```

## 4.2 Visualize annotation results
### 4.2.1 UMAP
```{r fig.width=5,fig.height=4}
Annotation.colours <- setNames(c("#83c5be", # CD4.T
                                 "#f5cac3", # CD8.T
                                 "#BC80BD", # NK.CD56hi
                                 "#E5C494", # NK.CD56lo
                                 "#C16610", # Prolif.lymph
                                 "#B3DE69", # ILC
                                 "#ffbe0b", # CD14.mono
                                 "#1F78B4", # CD16.mono
                                 "#66A61E", # cDC
                                 "#7400b8", # pDC
                                 "#748cab", # B
                                 "#1B9E77", # Plasmablast
                                 "#E7298A" # HSPC
                                 ),
               levels(seu$ann_lv1))

DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE,
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        #split.by = "Condition",
        group.by = "ann_lv1") + #NoLegend() +
  scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p1

suppressWarnings(p1)

```

```{r fig.width=6.5,fig.height=4}
Annotation2.colours <- c(c("#83c5be", # CD4.T
                                 "#f5cac3", # CD8.T
                                 "#BC80BD", # NK.CD56hi
                                 "#E5C494", # NK.CD56lo
                                 "#C16610", # Prolif.lymph
                                 "#B3DE69", # ILC
                                 "#ffbe0b", # CD14.mono
                                 "#1F78B4", # CD16.mono
                                 "#66A61E", # cDC
                                 "#7400b8", # pDC
                                 "#748cab", # B
                                 "#1B9E77", # Plasmablast
                                 "#E7298A" # HSPC
                                 ),
                              carto_pal(n=12,"Pastel")[1:11],
                              #carto_pal(n=12,"Antique")[1],carto_pal(n=12,"Antique")[3:4],
                              carto_pal(n=12,"Pastel")[12])
DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE,
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        #split.by = "Condition",
        group.by = "ann_lv2") + #NoLegend() +
  scale_color_manual(values=Annotation2.colours) +
  labs(x = NULL, y = NULL, title="PBMC") +
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p2

suppressWarnings(p2)

```

```{r fig.width=14.5,fig.height=4}
Annotation2.colours <- c(c("#83c5be", # CD4.T
                                 "#f5cac3", # CD8.T
                                 "#BC80BD", # NK.CD56hi
                                 "#E5C494", # NK.CD56lo
                                 "#C16610", # Prolif.lymph
                                 "#B3DE69", # ILC
                                 "#ffbe0b", # CD14.mono
                                 "#1F78B4", # CD16.mono
                                 "#66A61E", # cDC
                                 "#7400b8", # pDC
                                 "#748cab", # B
                                 "#1B9E77", # Plasmablast
                                 "#E7298A" # HSPC
                                 ),
                              carto_pal(n=12,"Pastel")[1:11],
                              #carto_pal(n=12,"Antique")[1],carto_pal(n=12,"Antique")[3:4],
                              carto_pal(n=12,"Pastel")[12])
DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE,
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        split.by = "Condition2",
        group.by = "ann_lv2") + #NoLegend() +
  scale_color_manual(values=Annotation2.colours) +
  labs(x = NULL, y = NULL, title=NULL) +
  guides(x = "none", y = "none") +
  theme(#plot.title.position = "panel",
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p2.5

suppressWarnings(p2.5)

```

### 4.2.2 Cell proportion

#### 4.2.2.1 Grouped by disease
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=3,fig.height=3.5}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p3 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.title = element_text(size=12),
        axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.text.y = element_text(size=10,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "PBMC cell-type proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank()) + NoLegend()
p3

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
md <- seu@meta.data
cells.keep <- rownames(md)[md$ann_lv1 %in% c("CD14.mono","CD16.mono","cDC","pDC")]
s.mye <- subset(seu, cells = cells.keep)
s.mye$ann_lv1 <- fct_drop(s.mye$ann_lv1)
s.mye$ann_lv2 <- fct_drop(s.mye$ann_lv2)


# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### T_NK {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
cells.keep <- rownames(md)[md$ann_lv1 %in% c("CD4T","CD8T","NK.CD56hi","NK.CD56lo","Prolif.lymph")]
s.Tlym <- subset(seu, cells = cells.keep)
s.Tlym$ann_lv1 <- fct_drop(s.Tlym$ann_lv1)
s.Tlym$ann_lv2 <- fct_drop(s.Tlym$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

#### 4.2.2.2 Grouped by sample
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.text.y = element_text(size=10,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  scale_x_discrete(labels=c(paste0("C-",1:5),paste0("CFN-",1:6),paste0("CFB-",1:6))) +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Lymphocyte {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

### 4.2.3 Cell type proportion changes [ann_lv1]
::::: {.panelset}

::: {.panel}
#### All cells {.panel-name}
```{r fig.width=10, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv1,
                             sample=seu$sampleID)
                             
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",65),rep("CF",156)),
                Condition2 = c(rep("C",65),rep("CFN",78),rep("CFB",78))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p

```

:::

::: {.panel}
#### Myeloid {.panel-name}
```{r fig.width=5, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$ID2)
                             
n <- length(levels(s.mye$ann_lv1))
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",5*n),rep("CF",12*n)),
                Condition2 = c(rep("C",5*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p3 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 12, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p3

```

:::

::: {.panel}
#### Lymphoid {.panel-name}
```{r fig.width=10, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$ID2)
                             
n <- length(levels(s.Tlym$ann_lv2))

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",5*n),rep("CF",12*n)),
                Condition2 = c(rep("C",5*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p4 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition3.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 12, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  ) 

p4

```

:::
:::::


### 4.2.4 Proportion CF vs C

#### 4.2.4.1 Grouped by disease
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=3,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv1,
                             sample=seu$Condition)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
s.mye <- subset(seu, ann_lv1 %in% c("CD14.mono","CD16.mono","cDC","pDC"))
s.mye$ann_lv1 <- fct_drop(s.mye$ann_lv1)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$Condition)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}

##### T_NK {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
s.Tlym <- subset(seu, ann_lv1 %in% c("CD4T","CD8T","NK.CD56hi","NK.CD56lo","Prolif.lymph"))
s.Tlym$ann_lv1 <- fct_drop(s.Tlym$ann_lv1)
s.Tlym$ann_lv2 <- fct_drop(s.Tlym$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$Condition)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

### 4.2.5 Cell type proportion changes [ann_lv2]
::::: {.panelset}

::: {.panel}
#### All cells {.panel-name}
```{r fig.width=10, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv1,
                             sample=seu$sampleID)
                             
n <- length(levels(seu$ann_lv1))
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",5*n),rep("CF",12*n)),
                Condition2 = c(rep("C",5*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition
  ) %>% 
  ungroup() %>%
  adjust_pvalue(p.col = "p", method = "BH") %>%
  add_significance("p.adj")
  


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p <- ggplot(df, aes(x = clusters, y = Freq, color = Condition)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p

```

:::

::: {.panel}
#### Myeloid {.panel-name}
```{r fig.width=6, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv2,
                             sample=s.mye$ID2)
                             
n <- length(levels(s.mye$ann_lv2))
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",5*n),rep("CF",12*n)),
                Condition2 = c(rep("C",5*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition
  ) %>% 
  ungroup() %>%
  adjust_pvalue(p.col = "p", method = "BH") %>%
  add_significance("p.adj")


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p5 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Proportion", x="PBMC myeloid cell type"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 10, colour = "black"),
    #axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.8, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 10)
  )

p5

```

:::

::: {.panel}
#### Lymphoid {.panel-name}
```{r fig.width=6, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$ID2)
                             
n <- length(levels(s.Tlym$ann_lv2))

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",5*n),rep("CF",12*n)),
                Condition2 = c(rep("C",5*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(Freq ~ Condition, ref.group = "C") %>%  # returns column `p`
  ungroup() %>%
  adjust_pvalue(p.col = "p", method = "BH") %>%
  add_significance("p.adj")


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p6 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Proportion", x="PBMC lymphoid cell type"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 10, colour = "black"),
    #axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.8, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 10)
  )

p6

```

:::
:::::

### 4.2.4 Plots of selected markers
```{r fig.width=8,fig.height=12}
# rearrange levels
#seu$ann_lv2 <- factor(seu$ann_lv2,levels=rev(levels(seu$ann_lv2)))

markers2 <- c("CD3D","CD3E","CCR7","TCF7","IL7R","LEF1","SELL",
              "CD4","CD40LG","CD8A","CD8B","CCL5","GZMK","KLRB1",
              "FOXP3","TIGIT","CTLA4",
              "TRDC1","TRGV2","TRDV1","GNLY","NCAM1",
              "MKI67","TOP2A",
              "PCDH9","TNFRSF18","TNFRSF4",
              "MS4A1","BANK1","CD79A","FCER2","TCL1A",
              "TNFRSF13B","AIM2","IGHM","CD27","CD44",
              "IGHA2","JCHAIN",
              "CD14","FCGR3A","CLEC10A","IL3RA","CD34","SPINK2") # Epithelial

# horizontal
p4 <- DotPlot(seu,
        features = rev(markers2),
        group.by = "ann_lv2",
        col.max = 2,
        col.min = 0,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="transparent", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) +
  labs(y = element_blank(), x = element_blank()) +
  theme(
        axis.text.x = element_text(size=10,
                                   face="plain",
                                   hjust = 1,vjust=1,angle=45),
        axis.text.y = element_text(size=10,face="italic"),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,angle=90),
        ) +
  scale_color_gradient(low = brewer.reds(9)[1], high=brewer.reds(9)[8], labels=c("0","","1","","2")) +
  #scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100),
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  coord_flip() +
  guides(
    color = guide_colorbar(
      title = "Mean Expression",
      title.position = "left",   # << here
      title.hjust = 0.5,
      barwidth = 1,
      barheight = 5
    ),
    size = guide_legend(
      override.aes = list(shape=21, colour="black", fill="#adb5bd"),
      title = "% of exp. cells",
      label.position = "right",
      title.position = "left",
      title.hjust = 0.5
    )
  )

# p2$guides$colour$title <- "Mean Expression"
# p2$guides$colour$title.position <- "top"
# p2$guides$colour$title.hjust <- 0.5
#seu$ann_lv2 <- factor(seu$ann_lv2,levels=rev(levels(seu$ann_lv2)))
suppressWarnings(print(p4))

```

```{r fig.width=8,fig.height=4}
# rearrange levels
seu$ann_lv1 <- factor(seu$ann_lv1,levels=rev(levels(seu$ann_lv1)))

p6 <- DotPlot(seu,
        features = c("CD3D","CD3E","CD40LG","CD4","CD8A","NCR1","GNLY","NCAM1","MKI67","TOP2A",
                     "TNFRSF18","TNFRSF4","CD14","FCGR3A","FCER1A","CLEC4A","IL3RA",
                     "MS4A1","CD79A","JCHAIN","CD34","SPINK2"),
        group.by = "ann_lv1",
        assay = "RNA",
        col.max = 2,
        col.min = 0,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="transparent", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) +
  labs(y = element_blank(), x = element_blank()) +
  theme(
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,angle=90),
        ) +
  scale_color_gradient(low = brewer.reds(9)[1], high=brewer.reds(9)[7]) +
  #scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100),
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(
    color = guide_colorbar(
      title = "Mean Expression",
      title.position = "left",   # << here
      title.hjust = 0.5,
      barwidth = 1,
      barheight = 5
    ),
    size = guide_legend(
      override.aes = list(shape=21, colour="black", fill="#adb5bd"),
      title = "% of exp. cells",
      label.position = "right",
      title.position = "left",
      title.hjust = 0.5
    )
  )

# p6$guides$colour$title <- "Mean Expression"
# p6$guides$colour$title.position <- "left"
# p6$guides$colour$title.hjust <- 0.5
seu$ann_lv1 <- factor(seu$ann_lv1,levels=rev(levels(seu$ann_lv1)))
suppressWarnings(print(p6))

```

# 5 Save objects
```{r}
if(!dir.exists(here("data","SCEs","analysis"))) {
  dir.create(here("data","SCEs","analysis"), recursive = TRUE)
}

out <- here("data","SCEs","analysis","pbmc.SEU.qs")
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

# out <- here("data","SCEs","manual.annot","pbmc.SEU.qs")
# if(!file.exists(out)) {
#   qsave(seu,file=out,nthreads=24)
# }
# 
# if(!dir.exists(here("data","SCEs","manual.annot"))) {
#   dir.create(here("data","SCEs","manual.annot"), recursive = TRUE)
# }
# 
# out <- here("data","SCEs","manual.annot","pbmc.SEU.qs")
# if(!file.exists(out)) {
#   qsave(seu,file=out,nthreads=24)
# }
# 
# out <- here("data","SCEs","manual.annot","pbmc_T.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("CD4.T","CD8.T"))
#   qsave(s,file=out,nthreads =24)
# }
# 
# out <- here("data","SCEs","manual.annot","pbmc_myeloid.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("CD14.mono","CD16.mono","cDC","pDC"))
#   qsave(s,file=out,nthreads =24)
# }
# 
# out <- here("data","SCEs","manual.annot","pbmc_mono.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("CD14.mono","CD16.mono"))
#   qsave(s,file=out,nthreads =24)
# }
# 
# out <- here("data","SCEs","manual.annot","pbmc_B.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("B","plasmablast"))
#   qsave(s,file=out,nthreads =24)
# }

```

# 6 Save figures
```{r}
celltype <- "combined"
if(!dir.exists(here("data","plots",celltype))) {
  dir.create(here("data","plots",celltype),recursive=TRUE)
}

# no split
# umap
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p1 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_ann_lv1.umap.jpeg"),
       height=4,
       width=5)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# umap
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p2 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_ann_lv2.umap.jpeg"),
       height=4,
       width=6.5)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# cellprop
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p3 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_prop.jpeg"),
       height=3.5,
       width=3)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)


# marker gene dotplot
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p4 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_marker_dotplot.jpeg"),
       height=10,
       width=8)
showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)


# cell type proportion test
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)
ggsave(plot=p5 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_myeloid_proptest.jpeg"),
       height=4,
       width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)


showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)
ggsave(plot=p6 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"pbmc_lymphoid_proptest.jpeg"),
       height=4,
       width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)


```


```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$init.clust,
                             sample=seu$sampleID)

props$Proportions %>% t() -> df
# 1) If `df` is a matrix or table, you can do:
df <- as.data.frame.matrix(df, stringsAsFactors = FALSE, check.names = FALSE)

# 2) (optional) If you want the rownames as a column called “sample”:
df$sampleID <- rownames(df)

# relocate sampleID to be the first column
df <- df %>%
  relocate(sampleID, .before = 1)

# 1) compute sample‐level mean
mean_mito <- seu@meta.data %>%
  as_tibble(rownames = "cell") %>%   # optional: drop rownames so they don’t interfere
  group_by(sampleID) %>%
  summarise(mean_mito_subset_percent = mean(subsets_Mito_percent, na.rm = TRUE))

# 2) join into your df
df <- df %>%
  left_join(mean_mito, by = "sampleID")

# add a batchID column by matching sampleID
df$batchID <- seu@meta.data$batchID[
  match(df$sampleID, seu@meta.data$sampleID)
]

# 1) count cells per sample
cell_counts <- seu@meta.data %>%
  as_tibble() %>%               # drop rownames
  count(sampleID, name = "nCells")

# 2) join into your df
df <- df %>%
  left_join(cell_counts, by = "sampleID")

df %>% write.csv("pbmc_cellProp_meta.csv",row.names = FALSE)
```
