---
title: 'BAL cell manual annotation'
author:
- name: Anson Wong
date: "`r Sys.Date()`"
description: null
output: html_document
editor_options:
  chunk_output_type: inline
---

```{r setup}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
xaringanExtra::use_panelset()

```

# 1 Load packages

```{r}
suppressPackageStartupMessages({
library(Seurat)
library(here)
library(qs)
library(forcats)
library(pals)
library(ggplot2)
library(speckle)
library(scCustomize)
library(patchwork)
library(dplyr)
library(Cepo)
library(rstatix)
library(ggpubr)
library(tibble)
library(readr)
library(rcartocolor)
})
options(future.globals.maxSize = 500000000000, future.seed=TRUE)
set.seed(1990)

```

# 2 Reference-based annotation
Reference-based annotation was performed on the clustered dataset using three references:
a. Wong et al 2025. For details, refer to analysis/05.annotation_Wong.Rmd

b. Azimuth PBMC using RunAzimuth(). Due to internal in-house server compatibility issues, 
R v4.2.2 instead of R v4.1.2 was used for running this step. 
or details, refer to code/azimuth_lungv2.R

## 2.1 Load object
```{r}
out <- here("data","SCEs","annotation",
            "bal.preprocessed.merged.integrated.clustered.Wong.Azimuth.rd.SEU.qs")
if (!file.exists(out)) {
  seu <- qread(here("data",
                    "SCEs",
                    "annotation",
                    "bal.preprocessed.merged.integrated.clustered.Wong.Azimuth.SEU.qs"), nthreads=24)
  
  DefaultAssay(seu) <- "SCT"
  # remove some info to reduce the file size
  seu@assays$prediction.score.ann_level_1 <- NULL
  seu@assays$prediction.score.ann_level_2 <- NULL
  seu@assays$prediction.score.ann_level_3 <- NULL
  seu@assays$prediction.score.ann_level_4 <- NULL
  seu@assays$prediction.score.ann_level_5 <- NULL
  seu@assays$prediction.score.ann_finest_level <- NULL
  seu@assays$prediction.score.celltype <- NULL
  seu@assays$prediction.score.sampleID <- NULL
  seu@reductions$azimuth.umap <- NULL
  seu@reductions$ref.pca <- NULL
  seu@reductions$ref.umap <- NULL
  seu@reductions$integrated_dr <- NULL
  seu@assays$refAssay <- NULL
  gc()
  
  # save object with reduced file size
  qsave(seu, file=out, nthreads=16)
  
} else {
  seu <- qread(out,nthreads=16)
}

Condition2.colours <- setNames(c("#26547c","#ffd166","#ef476f"),
                               c("C","CFN","CFB"))
```

## 2.2 Prepare object for downstream analysis
```{r}
seu.backup <- seu
seu$Condition <- fct_drop(seu$Condition)
seu$sampleID <- fct_drop(seu$sampleID)

# Assign ID for analysis
seu$sampleID <- factor(seu$sampleID,
                       levels=c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092", # healthy
                                "M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B", # CFN
                                "M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A"
                                ))

# Add Condition
Condition <- factor(
      case_when(
        seu$sampleID %in% c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092") ~ "C",
        seu$sampleID %in% c("M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B",
                            "M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A") ~ "CF"))

names(Condition) <- colnames(seu)
seu <- AddMetaData(seu, Condition, col.name = "Condition")
seu$Condition <- factor(seu$Condition,levels=c("C","CF"))

# Add Condition 3
Condition2 <- factor(
      case_when(
        seu$sampleID %in% c("M1N066","M1N075","M1N078","M1N080","M1N087","M1N092") ~ "C",
        seu$sampleID %in% c("M1C176C","M1C180D","M1C191B","M1C199B","M1C201A","M1C207B") ~ "CFN",
        seu$sampleID %in% c("M1C166(1)","M1C190C","M1C196B","M1C198B","M1C208A","M1C209A") ~ "CFB"))

names(Condition2) <- colnames(seu)
seu <- AddMetaData(seu, Condition2, col.name = "Condition2")
seu$Condition2 <- factor(seu$Condition2,levels=c("C","CFN","CFB"))

# Set ID for this study
ID <- setNames(c("C-NP1","C-NP2","C-NP3","C-NP4","C-4","C-5",
                 paste0("CF-",1:12)),
               levels(seu$sampleID))

ID <- setNames(factor(ID[seu$sampleID],
                      levels=c("C-NP1","C-NP2","C-NP3","C-NP4","C-4","C-5",
                               paste0("CF-",1:12))),
               colnames(seu))

seu <- AddMetaData(seu, ID, col.name = "ID")

# Set ID2 for this study
ID2 <- setNames(c("C-NP1","C-NP2","C-NP3","C-NP4","C-4","C-5",
                  paste0("CFN-",1:6),paste0("CFB-",1:6)),
               levels(seu$sampleID))

ID2 <- setNames(factor(ID2[seu$sampleID],
                       levels=c("C-NP1","C-NP2","C-NP3","C-NP4","C-4","C-5",
                                paste0("CFN-",1:6),paste0("CFB-",1:6))),
               colnames(seu))

seu <- AddMetaData(seu, ID2, col.name = "ID2")

# Re-order levels of SCT_snn_res.3
seu$SCT_snn_res.3 <- factor(seu$SCT_snn_res.3, levels=as.character(seq(1, length(levels(seu$SCT_snn_res.3)))))

# Number of droplets after annotation
dim(seu)
```

# 3. Visualize annotation
```{r}
# prepare colours
colours <- c(brewer.paired(12), brewer.pastel1(9),  brewer.pastel2(8), brewer.spectral(11), 
             rev(brewer.accent(8)), rev(brewer.dark2(8)), 
             brewer.set1(9), brewer.pastel2(8), brewer.spectral(11), 
             brewer.brbg(11)[1:5],rev(brewer.set3(12)),brewer.brbg(11)[6:11],
             brewer.piyg(11)[1:5],brewer.piyg(11)[7:11], brewer.piyg(11)[1:5],brewer.piyg(11)[7:11])

Condition.colours <- setNames(c("#81b29a","#e07a5f"),
                               levels(seu$Condition))

Condition2.colours <- setNames(c("#81b29a","#d69f7e","#ff8fab"),
                               levels(seu$Condition2))
```

## 3.1 UMAP 
### 3.1.1 Not split
```{r fig.width=8,fig.height=8}

group <- setNames(c("Wong_celltype",
                    "Azimuth_l3","Azimuth_finest",
                    "SCT_snn_res.3"),
                  c("predicted.celltype",
                    "predicted.ann_level_3","predicted.ann_finest_level",
                    "SCT_snn_res.3")
                  )
plots <- list()
plots <- sapply(names(group), FUN = function(x) {
  p <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = TRUE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        group.by = x) + NoLegend() +
  labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
  return(p)
})

((plots[[1]] + plots[[2]] + plot_layout(widths=c(1, 1)))/
    (plots[[3]] + plots[[4]] + plot_layout(widths=c(1, 1)))) +
  plot_layout(guides="collect")
```

### 3.1.2 SCT_snn_res.3 showing legend 
```{r fig.height=5, fig.width=6, fig.dpi=800}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        group.by = "SCT_snn_res.3") + #NoLegend() +
  #labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))

```


### 3.1.3 Split
```{r}
plots <- list()
plots <- sapply(names(group), FUN = function(x) {
  p <- DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, 
        cols=colours,
        split.by = "Condition2",
        group.by = x) + #NoLegend() +
  labs(x = NULL, y = NULL, title = group[x]) + 
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
  return(p)
})
```

::::: {.panelset}

::: {.panel}

#### Wong broad {.panel-name}

```{r fig.width=14,fig.height=4}
(plots[[1]]) + plot_layout(guides="collect")

```

:::

::: {.panel}

#### Azimuth {.panel-name}

```{r fig.width=18,fig.height=8}
(plots[[2]]/plots[[3]]) + plot_layout(guides="collect")

```

:::

::: {.panel}

#### SCT_snn_res.3 {.panel-name}

```{r fig.width=14,fig.height=4}
plots[[4]] + plot_layout(guides="collect")

```

:::

:::::

### 3.1.4 Check batch effect
::::: {.panelset}

::: {.panel}

#### By batch {.panel-name}
```{r fig.width=12,fig.height=4}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, ncol = 4,
        cols=colours,
        split.by = "batchID",
        group.by = "SCT_snn_res.3") +
  guides(x = "none", y = "none") +
  theme(
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
```
:::

::: {.panel}

#### By Condition {.panel-name}
```{r fig.width=12,fig.height=4}
DimPlot(seu, 
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE, 
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2, ncol = 5,
        cols=colours,
        split.by = "Condition2",
        group.by = "SCT_snn_res.3") + 
  guides(x = "none", y = "none") +
  theme(
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12))
```

:::
:::::

## 3.2 Proportion of cells per cluster
```{r}
labels <- group
p <- vector("list",length(labels))

for(label in names(labels)){
  seu@meta.data %>%
    ggplot(aes(x = !!sym(label), fill = !!sym(label))) +
    geom_bar() +
    geom_text(aes(label = ..count.., vjust=0.5, angle=45), size=3, stat = "count",
              colour = "black") +
    scale_y_log10() +
    theme(axis.text.x = element_blank(),
          axis.title.x = element_blank(),
          axis.ticks.x = element_blank()) +
    NoLegend() + scale_fill_manual(values = colours) +
    labs(y = "# cells (log scale)") -> p1
  
  seu@meta.data %>%
    dplyr::select(!!sym(label), Condition3) %>%
    group_by(!!sym(label), Condition3) %>%
    summarise(num = n()) %>%
    mutate(prop = num / sum(num)) %>%
  ggplot(aes(x = !!sym(label), y = prop * 100, 
             fill = Condition3)) + 
    geom_bar(stat = "identity") +
    theme(axis.text.x = element_text(angle = 45, size=6,
                                     vjust = 1, 
                                     hjust = 1)) +
    labs(y = "% Cells", fill = "Condition3") + scale_fill_manual(values = Condition3.colours)-> p2
  
  p1 / p2 -> p[[label]]
}

```

::::: {.panelset}

::: {.panel}

### Wong_broad {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.celltype + plot_layout(guides="collect")

```

:::

::: {.panel}

### Azimuth lv3 {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.ann_level_3 + plot_layout(guides="collect")

```

:::

::: {.panel}

### Azimuth finest {.panel-name}

```{r fig.width=8,fig.height=6}
p$predicted.ann_finest_level + plot_layout(guides="collect")

```

:::

::: {.panel}

### SCT_snn_res.3 {.panel-name}

```{r fig.width=8,fig.height=6}
p$SCT_snn_res.3 + plot_layout(guides="collect")

```

:::

:::::


## 3.3 Cepo
```{r}
# Run Cepo
cepoMarkers <- Cepo(seu[["SCT"]]@data,
                    seu$SCT_snn_res.3,
                    exprsPct = 0.2)
cepoMarkers$stats

sapply(1:ncol(cepoMarkers$stats), function(i){
  names(sort(cepoMarkers$stats[,i], decreasing = TRUE))[1:100]
}) %>% data.frame() -> dat

colnames(dat) <- gsub("^X","",colnames(cepoMarkers$stats))

# Get the column names of your data frame
current_colnames <- colnames(dat)

# Convert the names to numeric type, sort them, and then convert back to character
sorted_colnames <- as.character(sort(as.numeric(current_colnames)))

# Reorder the 'dat' data frame using the numerically sorted column names
dat <- dat[, sorted_colnames]

# Verify the new order
head(dat)

if (!dir.exists(here("data","Cepo"))) {
  dir.create(here("data","Cepo"))
}
dat %>% data.frame(check.names = FALSE) %>% write.csv(file = here("data","Cepo","bal.markers.SCT_snn_res.3.csv"),row.names = FALSE)

```

# 4. Cluster curation

## 4.1 DotPlots

::::: {.panelset}
::: {.panel}
### Broad marker genes {.panel-name}
```{r fig.width=18,fig.height=15}
# factorize metadata
seu$predicted.celltype <- factor(seu$predicted.celltype)
seu$predicted.ann_level_3 <- factor(seu$predicted.ann_level_3)
seu$predicted.ann_finest_level <- factor(seu$predicted.ann_finest_level)

# Gene list
## Wong 2025, derived from Fig1D
Wong.genes <- c("FABP4","GPD1","PCOLCE2","MME","SCD", # AM
                "VCAN","FCN1","EMP1","CD14","CCL2", # RecM
                "PCLAF","MKI67","TOP2A", # Prolif.M
                "CLEC10A","CD1C","LGALS2", # cDC
                "JCHAIN","PTPRS","CLEC4C", # pDC
                "CCR7","LAD1","COL19","CCL22","LAMP3","SLCO5A1", # DC migratory
                "CPA3","TPSAB1","TPSB2", # mast
                "CD3D","CD3G","CD3E", # T
                "CD4","FOXP3","TCF7","IL2RA","BCL6","CXCR5", # CD4
                "CD8A","CD8B","GZMB","GZMH","NKG7","KLRD1","EOMES", # CD4 & CD8
                "GNLY","NCAM1","KLRC1","NCR1","FCGR3A", # NK
                "MS4A1","EBF1","PAX5", # B
                "CHST9","MUC16","MUC4") # Epithelial

p2 <- DotPlot(seu,
        features = Wong.genes,
        group.by = "SCT_snn_res.3",
        col.max = 3,
        col.min = 0,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Ariel"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.bugn(9)[1], high=brewer.bugn(9)[8]) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p2$guides$colour$title <- "Mean Expression"
p2$guides$colour$title.position <- "top"
p2$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p2))

```

:::

::: {.panel}
#### Macrophage {.panel-name}

```{r fig.width=10,fig.height=15}
s <- subset(seu, SCT_snn_res.3 %in% c("42","17","43","52","6","8","23","37","41","54",
                                      "53","50","25"),invert=TRUE)
s$SCT_snn_res.3 <- fct_drop(s$SCT_snn_res.3)

Wong.macro.genes <- c("CD14","FCGR3A","IL1B","S100A12","MARCKS",
                      "F13A1","MAF","FOLR2","CCR2","VCAN",
                      "PLA2G7","SPP1",
                      "CCL2","CCL8","CXCL9","ISG15","GBP4","GBP5","IDO1","GCH1",
                      "ITGAM","FCGR1A","CSF1R",
                      "C1QA","FABP4","CD163","MARCO","GPD1","PCOLCE2","MME","SCD",
                      "MT1X","MT1G","MT1F","IFI27","CCL18","APOC2","ABCA1","CD36","LAMP1")
p2 <- DotPlot(s,
        features = Wong.macro.genes,
        group.by = "SCT_snn_res.3",
        col.max = 2,
        col.min = 0,
        dot.scale = 5
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="black", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) + 
  labs(y = element_blank(), x = element_blank()) + 
  theme(text=element_text(family="Ariel"),
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="horizontal",
        legend.position="bottom",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,family = "Arial"),
        ) +
  scale_color_gradient(low = brewer.bugn(9)[1], high=brewer.bugn(9)[8]) +
  scale_size_continuous(limits = c(0, 100), 
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(size=guide_legend(override.aes=list(shape=21, 
                                             colour="black", 
                                             fill="#adb5bd"),
                                             title="% of exp. cells",
                                             label.position="bottom",
                                             title.position="top",
                                             title.hjust=0.5
                                             ))

p2$guides$colour$title <- "Mean Expression"
p2$guides$colour$title.position <- "top"
p2$guides$colour$title.hjust <- 0.5
suppressWarnings(print(p2))

```

:::

:::::

## 4.2 Check predictive score
::::: {.panelset}

::: {.panel}

### Azimuth {.panel-name}
```{r fig.width=16,fig.height=12}
# Plot Azimuth predictive score
seu@meta.data %>%
  ggplot(aes(y = predicted.ann_level_3.score,
             x = predicted.ann_level_3,
             fill = predicted.ann_level_3)) +
  facet_wrap(~SCT_snn_res.3, ncol = 6) +
  geom_violin(scale = "width") +
  scale_fill_manual(values = colours) +
  theme(text = element_text(size = 8),
        axis.text.x = element_text(angle = 90,
                            vjust = 0.5,
                            hjust = 1))
```

:::

::: {.panel}

### Wong broad {.panel-name}
```{r fig.width=16,fig.height=12}
# Plot Azimuth predictive score
seu@meta.data %>%
  ggplot(aes(y = predicted.celltype.score,
             x = predicted.celltype,
             fill = predicted.celltype)) +
  facet_wrap(~SCT_snn_res.3, ncol = 6) +
  geom_violin(scale = "width") +
  scale_fill_manual(values = colours) +
  theme(text = element_text(size = 8),
        axis.text.x = element_text(angle = 90,
                            vjust = 0.5,
                            hjust = 1))
```

:::

:::::

# 5. Manual annotation
## 5.1 Annotation
<!-- read from prevoius file -->
<!-- ```{r} -->
<!-- # 1) Read annotation file (all samples) -->
<!-- ann_path <- "/group/canc2/anson/working/cf-eti-bal/data/annotation/bal_annotations.csv" -->
<!-- ann_lv1  <- read_csv(ann_path, show_col_types = FALSE) -->

<!-- cells_seu <- rownames(seu@meta.data) -->

<!-- # Create a named vector: names = cell barcodes, values = subtype -->
<!-- ann_lv1_vec <- ann_lv1 %>% -->
<!--   distinct(cell, .keep_all = TRUE) %>% -->
<!--   filter(cell %in% cells_seu) %>% -->
<!--   { setNames(as.character(.$subtype), .$cell) } -->

<!-- # AddMetaData will match by names(ann_lv1_vec) to colnames(seu) -->
<!-- seu <- Seurat::AddMetaData(seu, metadata = ann_lv1_vec, col.name = "ann_lv1") -->

<!-- # check -->
<!-- table(is.na(seu$ann_lv1), useNA = "ifany") -->
<!-- head(seu$ann_lv1) -->

<!-- ``` -->

### 5.1.1 Level 1
```{r}
# Use Azimuth result to annotate mast cells
draft1 <- factor(
      case_when(
        seu$predicted.celltype == "Mast" ~ "Mast"
        ))

seu <- AddMetaData(seu, draft1, col.name = "draft1")

# v250321
draft2 <- setNames(c("AM","AM","AM","AM","AM", # 1-5
                     "T/NK","AM","T/NK","AM","AM", # 6-10
                     "AM","AM","AM","RecM","AM", # 11-15
                     "RecM","B","AM","AM","RecM", # 16-20
                     "AM","AM","T/NK","AM","cDC", # 21-25
                     "AM","RecM","AM","AM","AM", # 26-30
                     "AM","AM","AM","AM","AM", # 31-35
                     "Prolif.M","T/NK","RecM","AM","AM", # 36-40
                     "T/NK","pDC","Epithelial","AM","Prolif.M", # 41-45
                     "RecM","AM","AM","RecM","cDC", # 46-50
                     "AM","Epithelial","cDC","T/NK" # 51-54
                     ), levels(seu$SCT_snn_res.3))

draft2 <- factor(setNames(draft2[seu$SCT_snn_res.3],colnames(seu)))

seu <- AddMetaData(seu, draft2, col.name = "draft2")

# coalesce two draft columns
seu@meta.data %>%
  mutate(ann_lv1=coalesce(draft1,draft2)) -> seu@meta.data

# reorder levels
seu$ann_lv1 <- factor(seu$ann_lv1,
                       levels=c("AM","RecM","Prolif.M","cDC","pDC",
                                "T/NK","B","Mast","Epithelial"))

seu$draft1 <- NULL
seu$draft2 <- NULL

# Number of cells per celltype
summary(seu$ann_lv1)
```

### 5.1.2 Level 2
```{r}
# Use Azimuth result to annotate mast cells
draft1 <- factor(
      case_when(
        seu$predicted.celltype == "Mast" ~ "Mast"
        ))

seu <- AddMetaData(seu, draft1, col.name = "draft1")

# v251215
draft2 <- setNames(c("AM-IFI27hi","AM","AM-CCL","AM","AM-IFI27hi", # 1-5
                     "T/NK","AM","T/NK","AM-lipid","AM", # 6-10
                     "AM","AM","AM","IM","AM", # 11-15
                     "CD14.mono","B","AM","AM","MDM-PLA2G7hi", # 16-20
                     "AM","AM","T/NK","AM","cDC", # 21-25
                     "AM-IFI27hi","MDM-FABP4lo","AM","AM","AM-IFI27hi", # 26-30
                     "AM-MT","AM-IFI27hi","AM","AM","AM", # 31-35
                     "Prolif.M","T/NK","MDM-IV","AM","AM", # 36-40
                     "T/NK","pDC","Epithelial","AM","Prolif.M", # 41-45
                     "MDM-inflam","AM","AM-IFI27hi","IM","DC.migratory", # 46-50
                     "AM","Epithelial","cDC","T/NK" # 51-54
                     ), levels(seu$SCT_snn_res.3))

draft2 <- factor(setNames(draft2[seu$SCT_snn_res.3],colnames(seu)))

seu <- AddMetaData(seu, draft2, col.name = "draft2")

# coalesce two draft columns
seu@meta.data %>%
  mutate(ann_lv2=coalesce(draft1,draft2)) -> seu@meta.data


# BAL cell names (must be length 110809)
cells <- rownames(seu@meta.data)  # same as colnames(seu)

t_ann <- readr::read_csv(
  here::here("data","annotation","balTcell_subtype_annotations.csv"),
  show_col_types = FALSE
) %>%
  distinct(cell, .keep_all = TRUE) %>%
  mutate(subtype = as.character(subtype))  # keep "Ambiguous" as-is

t_map <- setNames(t_ann$subtype, t_ann$cell)

ann_lv2_old <- as.character(seu$ann_lv2)

t_override <- dplyr::case_when(
  seu$ann_lv1 == "T/NK" & cells %in% names(t_map) ~ unname(t_map[cells]),
  seu$ann_lv1 == "T/NK" & !(cells %in% names(t_map)) ~ "Ambiguous",
  TRUE ~ NA_character_
)

seu$ann_lv2 <- dplyr::coalesce(t_override, ann_lv2_old)

# quick checks
table(seu$ann_lv1 == "T/NK", is.na(t_override))
table(seu$ann_lv2[seu$ann_lv1 == "T/NK"], useNA = "ifany")

# Factorize (include Ambiguous level)
seu$ann_lv2 <- factor(
  seu$ann_lv2,
  levels = c(
    "AM","AM-CCL","AM-MT","AM-lipid","AM-IFI27hi",
    "CD14.mono","MDM-IV","MDM-FABP4lo","MDM-inflam","MDM-PLA2G7hi","IM","Prolif.M",
    "cDC","pDC","DC.migratory","Mast",
    "CD4T.naive","CD4T.memory","CD4T.TRM","Treg",
    "CD8T.EM","CD8T.TRM",
    "NKT","NK","ILC","Prolif.T/NK",
    "B","Epithelial","Ambiguous"
  )
)

summary(seu$ann_lv2)

```

### 5.1.3 Refine ann_lv1
```{r}
## Refine ann_lv1 for cells currently annotated as T/NK using ann_lv2 subtypes
## Rules:
## - ann_lv2 == "NK"            -> ann_lv1 = "NK"
## - ann_lv2 == "Treg"          -> ann_lv1 = "Treg"
## - ann_lv2 starts with "CD4"  -> ann_lv1 = "CD4T"
## - ann_lv2 starts with "CD8"  -> ann_lv1 = "CD8T"
## - ann_lv2 == "Ambiguous"     -> ann_lv1 = "Ambiguous"
## - otherwise keep existing ann_lv1

seu@meta.data <- seu@meta.data %>%
  dplyr::mutate(
    ann_lv1 = dplyr::case_when(
      ann_lv1 == "T/NK" & ann_lv2 == "NK" ~ "NK",
      ann_lv1 == "T/NK" & ann_lv2 == "Treg" ~ "Treg",
      ann_lv1 == "T/NK" & ann_lv2 == "NKT" ~ "NKT",
      ann_lv1 == "T/NK" & ann_lv2 == "ILC" ~ "ILC",
      ann_lv1 == "T/NK" & ann_lv2 == "Prolif.T/NK" ~ "Prolif.T/NK",
      ann_lv1 == "T/NK" & stringr::str_starts(as.character(ann_lv2), "CD4") ~ "CD4T",
      ann_lv1 == "T/NK" & stringr::str_starts(as.character(ann_lv2), "CD8") ~ "CD8T",
      ann_lv1 == "T/NK" & ann_lv2 == "Ambiguous" ~ "Ambiguous",
      TRUE ~ as.character(ann_lv1)
    ),
    ann_lv1 = factor(ann_lv1)
  )

# Quick check: remaining un-resolved T/NK (if any)
table(seu$ann_lv1, useNA = "ifany")

# Optional: set a preferred order for ann_lv1 levels (edit as you like)
seu$ann_lv1 <- factor(
  seu$ann_lv1,
  levels = c("AM","RecM","Prolif.M","cDC","pDC",
             "CD4T","CD8T","Treg","NKT","NK","ILC","Prolif.T/NK",
             "B","Mast","Epithelial","Ambiguous")
)

```

## 5.2 UMAP
### 5.2.1 ann_lv1
```{r fig.width=5,fig.height=4}

Annotation.colours <- setNames(c("#DD927E", # AM
                                 "#C5A07A", # RecM
                                 "#7f5539", # Prolif M
                                 "#979EC1", # cDC
                                 "#DB98AE", # pDC
                                 #"#BE93C6", # Migratory.DC
                                 "#a3b18a", # CD4T
                                 "#588157", # CD8T
                                 "#054a91", # Treg
                                 "#432818", # NKT
                                 "#BE93C6", # NK
                                 "#e27396", # ILC
                                 "#ff6700", # Prolif T/NK
                                 "#f6bd60", # B
                                 "#5a189a", # Mast
                                 "#b23a48", # Epithelial
                                 "#A5AA99" # Ambiguous
                                 ),
               levels(seu$ann_lv1))

DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = FALSE,
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        #split.by = "Condition",
        group.by = "ann_lv1") + #NoLegend() +
  scale_color_manual(values=Annotation.colours) +
  labs(x = NULL, y = NULL) +
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        text=element_text(family="Arial"),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p1

suppressWarnings(p1)

```

### 5.2.2 ann_lv2
```{r fig.width=6.5,fig.height=4}

Annotation2.colours <- c("#83c5be", # CD4.T
                                 "#f5cac3", # CD8.T
                                 "#BC80BD", # NK.CD56hi
                                 "#E5C494", # NK.CD56lo
                                 "#C16610", # Prolif.lymph
                                 "#B3DE69", # ILC
                                 "#ffbe0b", # CD14.mono
                                 "#1F78B4", # CD16.mono
                                 "#66A61E", # cDC
                                 "#7400b8", # pDC
                                 "#748cab", # B
                                 "#1B9E77", # Plasmablast
                                 "#E7298A" # HSPC
                                 ,
                              carto_pal(n=12,"Pastel")[1:11],
                              carto_pal(n=12,"Antique")[1],carto_pal(n=12,"Antique")[3:5],
                              carto_pal(n=12,"Pastel")[12])

DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE,
        shuffle = FALSE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        #split.by = "Condition",
        group.by = "ann_lv2") + #NoLegend() +
  scale_color_manual(values=Annotation2.colours) +
  labs(x = NULL, y = NULL, title="BAL") +
  guides(x = "none", y = "none") +
  theme(#plot.title = element_blank(),
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p2

suppressWarnings(p2)

```

split
```{r fig.width=14.5,fig.height=4}

DimPlot(seu,
        reduction = 'umap',
        pt.size = 0.1,
        label.box = FALSE,
        label = TRUE,
        shuffle = TRUE,
        raster=FALSE,
        repel = TRUE,
        label.size = 2,
        split.by = "Condition2",
        group.by = "ann_lv2") + #NoLegend() +
  scale_color_manual(values=Annotation2.colours) +
  labs(x = NULL, y = NULL, title=NULL) +
  guides(x = "none", y = "none") +
  theme(#plot.title.position = "panel",
        axis.text = element_text(size=8),
        axis.title.y=element_text(vjust=-1,
                                  size=10),
        axis.title.x=element_text(#vjust=10,
                                  size=10),
        legend.text = element_text(size = 10),
        strip.text.x = element_text(face = "bold", vjust=2, size=12)) -> p2.5

suppressWarnings(p2.5)

```

## 5.3 Cell proportion [ann_lv1]
### 5.3.1 Grouped by disease
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=3,fig.height=3.5}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p3 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.title = element_text(size=12),
        axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.text.y = element_text(size=10,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "BAL cell-type proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank()) + NoLegend()
p3

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
md <- seu@meta.data
cells.keep <- rownames(md)[md$ann_lv1 %in% c("AM","RecM","Prolif.M","cDC","pDC","Mast")]
s.mye <- subset(seu, cells = cells.keep)
s.mye$ann_lv1 <- fct_drop(s.mye$ann_lv1)
s.mye$ann_lv2 <- fct_drop(s.mye$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### T_NK {.panel-name}
```{r fig.width=3,fig.height=4}
# subset object
cells.keep <- rownames(md)[md$ann_lv1 %in% c("CD4T","CD8T","Treg","NKT","NK","ILC","Prolif.T/NK","B")]
s.Tlym <- subset(seu, cells = cells.keep)
s.Tlym$ann_lv1 <- fct_drop(s.Tlym$ann_lv1)
s.Tlym$ann_lv2 <- fct_drop(s.Tlym$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv1,
                             sample=s.Tlym$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

### 5.3.2 Grouped by sample
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv1,
                             sample=seu$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.text.y = element_text(size=10,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  scale_x_discrete(labels=c(paste0("C-",1:5),paste0("CFN-",1:6),paste0("CFB-",1:6))) +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Lymphocyte {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv1,
                             sample=s.Tlym$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::



## 5.4 Cell proportion [ann_lv2]
### 5.4.1 Grouped by disease
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=5,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=4,fig.height=4}
s.mye$ann_lv2 <- fct_drop(s.mye$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv2,
                             sample=s.mye$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### T_NK {.panel-name}
```{r fig.width=3,fig.height=4}
s.Tlym$ann_lv2 <- fct_drop(s.Tlym$ann_lv2)

# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$Condition2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

### 5.3.2 Grouped by sample
::::: {.panelset}

::: {.panel}
##### All cells {.panel-name}
```{r fig.width=9,fig.height=3}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=10,
                                   colour = "black"),
        axis.text.y = element_text(size=10,
                                   colour = "black"),
        legend.text = element_text(size = 10),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  scale_x_discrete(labels=c(paste0("C-",1:6),paste0("CFN-",1:6),paste0("CFB-",1:6))) +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Myeloid {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv2,
                             sample=s.mye$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::

::: {.panel}
##### Lymphocyte {.panel-name}
```{r fig.width=8,fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$ID2)

props$Proportions %>%
  data.frame() -> df

# stacked bar plot of cell proportion per sampleID
p2 <- ggplot(df, aes(x = sample, y = Freq, fill = clusters)) +
  geom_bar(stat = "identity", position="stack") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,
                                   vjust = 1,
                                   hjust = 1,
                                   size=8,
                                   colour = "black"),
        axis.text.y = element_text(size=8,
                                   colour = "black"),
        legend.text = element_text(size = 8),
        legend.title = element_text(face="bold", size=12)) +
  labs( y = "Proportion", fill="Condition", x=NULL) + #NoLegend() +
  guides(fill = guide_legend(keyheight=0.8,
                             keywidth=0.8)) +
  scale_fill_manual(values = Annotation2.colours) +
  theme(legend.text.align = 0, legend.title = element_blank())
p2

```
:::
:::::

# 5.4 Cell type proportion changes
### 5.4.1 ann_lv1
::::: {.panelset}

::: {.panel}
#### All cells {.panel-name}
```{r fig.width=10, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv1,
                             sample=seu$sampleID)

n <- nlevels(seu$ann_lv1)

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",n*6),rep("CF",n*12)),
                Condition2 = c(rep("C",n*6),rep("CFN",n*6),rep("CFB",n*6))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p

```

:::

::: {.panel}
#### Myeloid {.panel-name}
```{r fig.width=5, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv1,
                             sample=s.mye$ID2)
                             
n <- length(levels(s.mye$ann_lv1))
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",6*n),rep("CF",12*n)),
                Condition2 = c(rep("C",6*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p3 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 12, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p3

```

:::

::: {.panel}
#### Lymphoid {.panel-name}
```{r fig.width=5, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv1,
                             sample=s.Tlym$ID2)
                             
n <- length(levels(s.Tlym$ann_lv1))

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",6*n),rep("CF",12*n)),
                Condition2 = c(rep("C",6*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p4 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 12, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  ) 

p4

```

:::
:::::


### 5.4.2 ann_lv2
::::: {.panelset}

::: {.panel}
#### All cells {.panel-name}
```{r fig.width=10, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=seu$ann_lv2,
                             sample=seu$sampleID)

n <- nlevels(seu$ann_lv2)

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",n*6),rep("CF",n*12)),
                Condition2 = c(rep("C",n*6),rep("CFN",n*6),rep("CFB",n*6))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Cell proportion"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 12, colour = "black"),
    axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.75, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 12)
  )

p

```

:::

::: {.panel}
#### Myeloid {.panel-name}
```{r fig.width=6, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.mye$ann_lv2,
                             sample=s.mye$ID2)
                             
n <- length(levels(s.mye$ann_lv2))
props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",6*n),rep("CF",12*n)),
                Condition2 = c(rep("C",6*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  dplyr::filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p5 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Proportion", x="BAL myeloid cell type"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 10, colour = "black"),
    #axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.8, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 10)
  )

p5

```

:::

::: {.panel}
#### Lymphoid {.panel-name}
```{r fig.width=6, fig.height=4}
# calculate the proportion of cells --------------------------------------------
props <- getTransformedProps(clusters=s.Tlym$ann_lv2,
                             sample=s.Tlym$ID2)
                             
n <- length(levels(s.Tlym$ann_lv2))

props$Proportions %>%
  data.frame %>%
  dplyr::mutate(Condition = c(rep("C",6*n),rep("CF",12*n)),
                Condition2 = c(rep("C",6*n),rep("CFN",6*n),rep("CFB",6*n))) -> df

df$Condition <- factor(df$Condition, levels=c("C","CF"))
df$Condition2 <- factor(df$Condition2, levels=c("C","CFB","CFN"))

# Wilcox test on pairwise comparisons of control, CF mild and CF severe --------
stat.test <- df %>%
  group_by(clusters) %>%
  wilcox_test(
    ref.group = "C",
    Freq ~ Condition2,
    p.adjust.method = "BH"
  ) %>%
  filter(p.adj < 0.1)


stat.test <- stat.test %>%
  add_xy_position(x = "clusters", dodge = 0.8)

# box plot
dodge_width <- 0.8  # adjust this value to widen spacing between groups within clusters

p6 <- ggplot(df, aes(x = clusters, y = Freq, color = Condition2)) +
  geom_boxplot(
    fill = NA,
    outlier.shape = NA,
    position = position_dodge(width = dodge_width)
  ) +
  geom_point(
    position = position_dodge(width = dodge_width),
    size = 2, alpha = 0.8
  ) +
  stat_pvalue_manual(
    stat.test,
    label = "p.adj.signif",
    tip.length = 0.01,
    bracket.size = 0.4,
    hide.ns = TRUE
  ) +
  scale_color_manual(values = Condition2.colours) +
  labs(
    y = "Proportion", x="BAL lymphoid cell type"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(angle = 40, hjust = 1, size = 10, colour = "black"),
    axis.text.y = element_text(size = 10, colour = "black"),
    #axis.title.x = element_blank(),
    legend.title = element_blank(),
    legend.position = c(0.8, 0.99),
    legend.justification = c(0, 1),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key.size = unit(1, "lines"),
    legend.text = element_text(size = 10)
  )

p6

```

:::
:::::

### 5.2.4 Plots of selected markers
```{r fig.width=10,fig.height=4}
# rearrange levels
seu$ann_lv1 <- factor(seu$ann_lv1,levels=rev(levels(seu$ann_lv1)))

Wong.genes <- c("FABP4","GPD1","PCOLCE2","MME","SCD", # AM
                "VCAN","FCN1","CCL2", # RecM
                "PCLAF","MKI67","TOP2A", # Prolif.M
                "CLEC10A","CD1C","LGALS2", # cDC
                "JCHAIN","PTPRS","CLEC4C", # pDC
                "CD3D","CD3G","CD40LG","CD8A","FOXP3", # CD4.T
                "NCAM1","KLRC1", # NK
                "MS4A1","EBF1","PAX5", # B
                "CPA3","TPSAB1","TPSB2", # mast
                "CHST9","MUC16","MUC4") # Epithelial

# horizontal
p2 <- DotPlot(seu,
        features = Wong.genes,
        group.by = "ann_lv1",
        col.max = 2,
        col.min = 0,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="transparent", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) +
  labs(y = element_blank(), x = element_blank()) +
  theme(
        axis.text.x = element_text(size=10,
                                   face="italic",
                                   angle = 45,
                                   hjust = 1,
                                   vjust = 1),
        axis.text.y = element_text(size=10),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,angle=90),
        ) +
  scale_color_gradient(low = brewer.reds(9)[1], high=brewer.reds(9)[8], labels=c("0","","1","","2")) +
  #scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100),
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  guides(
    color = guide_colorbar(
      title = "Mean Expression",
      title.position = "left",   # << here
      title.hjust = 0.5,
      barwidth = 1,
      barheight = 5
    ),
    size = guide_legend(
      override.aes = list(shape=21, colour="black", fill="#adb5bd"),
      title = "% of exp. cells",
      label.position = "right",
      title.position = "left",
      title.hjust = 0.5
    )
  )

# p2$guides$colour$title <- "Mean Expression"
# p2$guides$colour$title.position <- "top"
# p2$guides$colour$title.hjust <- 0.5
seu$ann_lv1 <- factor(seu$ann_lv1,levels=rev(levels(seu$ann_lv1)))
suppressWarnings(print(p2))

```

### 5.2.4 Plots of selected markers
```{r fig.width=10,fig.height=15}
# rearrange levels
#seu$ann_lv2 <- factor(seu$ann_lv2,levels=rev(levels(seu$ann_lv2)))

markers2 <- c("GPD1","RBP4","CD14","FABP4","CCL4L2","CCL20","MT1G","MT1F","MT1X", # AM
              "CD36","NR1H3","IFI27","CCR2","FCN1",
              "FOLR2","F13A1","RNASE1","VCAN","CCL2","IL1B","CD300E",
              "CXCL9","CXCL10","CXCL11",
              "PCLAF","MKI67","TOP2A", # Prolif.M
              "CLEC10A","CD1C","LGALS2", # cDC
              "JCHAIN","PTPRS","CLEC4C", # pDC
              "CCR7","CCL22", #DC migratory
              "CPA3","TPSAB1","TPSB2", # mast
              "CD3D","CD3E","TRAC","TRDC","CD4","CD40LG","CD8A","CD8B",
              "LEF1","ZNF683","IFNG","GZMK","GZMA","FOXP3","CTLA4","TIGIT",
              "ITGA1","ITGAE","CD69",
              "NCAM1","KLRF1","GNLY", # NK
              "KIT","PCDH9", #ILC
              "MS4A1","EBF1","PAX5", # B
              "CHST9","MUC16","MUC4") # Epithelial

# horizontal
p4 <- DotPlot(seu,
        features = rev(markers2),
        group.by = "ann_lv2",
        col.max = 2,
        col.min = 0,
        dot.scale = 6
        ) + #coord_flip() +
  geom_point(aes(size=pct.exp), shape = 21, colour="transparent", stroke=0.5) +
  FontSize(y.text = 10, x.text = 8) +
  labs(y = element_blank(), x = element_blank()) +
  theme(
        axis.text.x = element_text(size=10,
                                   face="plain",
                                   hjust = 1,vjust=1,angle=45),
        axis.text.y = element_text(size=10,face="italic"),
        legend.direction="vertical",
        legend.position="right",
        legend.justification = "right",
        legend.box="vertical",
        legend.text = element_text(size = 8),
        legend.title = element_text(size = 8,angle=90),
        ) +
  scale_color_gradient(low = brewer.reds(9)[1], high=brewer.reds(9)[8], labels=c("0","","1","","2")) +
  #scale_y_discrete(labels = subtype.labels) +
  scale_size_continuous(limits = c(0, 100),
                        breaks = c(0, 25, 50, 75, 100),
                        labels=c(0,"",50,"",100)) +
  coord_flip() +
  guides(
    color = guide_colorbar(
      title = "Mean Expression",
      title.position = "left",   # << here
      title.hjust = 0.5,
      barwidth = 1,
      barheight = 5
    ),
    size = guide_legend(
      override.aes = list(shape=21, colour="black", fill="#adb5bd"),
      title = "% of exp. cells",
      label.position = "right",
      title.position = "left",
      title.hjust = 0.5
    )
  )

# p2$guides$colour$title <- "Mean Expression"
# p2$guides$colour$title.position <- "top"
# p2$guides$colour$title.hjust <- 0.5
#seu$ann_lv2 <- factor(seu$ann_lv2,levels=rev(levels(seu$ann_lv2)))
suppressWarnings(print(p4))

```

# 6 Save objects
```{r}
if(!dir.exists(here("data","SCEs","analysis"))) {
  dir.create(here("data","SCEs","analysis"), recursive = TRUE)
}

out <- here("data","SCEs","analysis","bal.SEU.qs")
if(!file.exists(out)) {
  qsave(seu,file=out,nthreads=24)
}

# out <- here("data","SCEs","manual.annot","bal_RecM.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("RecM"))
#   s$init.clust <- fct_drop(s$init.clust)
#   qsave(s,file=out,nthreads =16)
# }
# 
# out <- here("data","SCEs","manual.annot","bal_myeloid.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("RecM","cDC","pDC"))
#   s$init.clust <- fct_drop(s$init.clust)
#   qsave(s,file=out,nthreads =16)
# }
# 
# out <- here("data","SCEs","manual.annot","bal_T.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("T/NK"))
#   s$init.clust <- fct_drop(s$init.clust)
#   qsave(s,file=out,nthreads =16)
# }
# 
# out <- here("data","SCEs","manual.annot","bal_AM.SEU.qs")
# if(!file.exists(out)) {
#   s <- subset(seu, init.clust %in% c("AM","Prolif.M"))
#   s$init.clust <- fct_drop(s$init.clust)
#   qsave(s,file=out,nthreads =16)
# }
```

# 7 Save figures
```{r}
celltype <- "combined"
if(!dir.exists(here("data","plots",celltype))) {
  dir.create(here("data","plots",celltype),recursive=TRUE)
}

# no split
# umap
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p2 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"bal_ann_lv2.umap.jpeg"),
       height=4,
       width=6.5)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# cellprop
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p3 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"bal_prop.jpeg"),
       height=3.5,
       width=3)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# marker gene dotplot
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)

ggsave(plot=p4 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"bal_marker_dotplot.jpeg"),
       height=15,
       width=10)
showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# cell type proportion test
showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)
ggsave(plot=p5 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"bal_myeloid_proptest.jpeg"),
       height=4,
       width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)


showtext::showtext_auto()
showtext::showtext_opts(dpi = 1600)
ggsave(plot=p6 + theme(plot.tag = element_text(size = 14, face = "bold")),
       device="jpeg",
       dpi=1600,
       filename = here("data","plots",celltype,"bal_lymphoid_proptest.jpeg"),
       height=4,
       width=6)

showtext::showtext_auto()
showtext::showtext_opts(dpi = 100)

# # cellprop test
# showtext::showtext_auto()
# showtext::showtext_opts(dpi = 1600)
#
# p3 +
#   theme(plot.tag = element_text(size = 14, face = "bold"))
#
# ggsave(plot=last_plot(),
#        device="jpeg",
#        dpi=1600,
#        filename = here("data","plots",celltype,"pbmc_myeloid_proptest.jpeg"),
#        height=4,
#        width=5)
#



```